---
title: "OAI - Data Analysis"
author: "Mitchell Henschel"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: true
    theme: journal
---

## Settings

```{r}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(tidyverse)
options(scipen = 999) # No scientific notation
```

### Data Loading

```{r loadData, message=FALSE, warning=FALSE}
DATAPATH <- Sys.getenv("OAI_DATA")
if (DATAPATH == "" ) stop( "Please add datapath to OAI_DATA" )

source("OAI_LoadData.R", chdir = T) # LoadData file must be in working directory

if(!exists('data_full')) {
  
  if(!exists('bsln')) {
    bsln <- getBaselineData(DATAPATH)
  }
  
  data_full <- getCompleteData(DATAPATH, cluster = "K.5.Clusters")
  bsln$EVNT <- data_full$EVNT
  
  remove(data_full)
}
```

#### Missing Data:

-   There are `r sum(!complete.cases(bsln))` entries in with missing data.
-   `r sum(!complete.cases(bsln[as.numeric(bsln$EVNT) >= 4,]))` who received knee replacements
-   `r sum(!complete.cases(bsln[as.numeric(bsln$EVNT) < 4,]))` who did not

These observations have at least one missing value

```{r omitNa}
data_missing <- bsln[rowSums(is.na(bsln)) > 0,]
# bsln <- na.omit(bsln)
```

There are 24 variables of interest in this data set:

-   ID

-   [AGE](#age)

-   [SEX](#sex) (1 = Male, 2 = Female)

-   [Race](#race) (2 factor columns)

    -   RACE_AA - African-American
    -   RACE_NW - Other Non-White
    -   White (reference)

-   [ETHNICITY](#ethnicity) (factor) - Reported Hispanic or latino

-   [PASE](#pase) - Physical Activity Score for the Elderly

-   [MEDINS](#medins) - Insurance prescript payment (factor)

-   [WOMADL](#womadl) - WOMAC Disability Score

-   [WOMKP](#womkp) - WOMAC Pain Score

-   [WOMSTF](#womstf) - WOMAAC Stiffness Score

-   [BMI](#bmi)

-   [HEIGHT](#height) - Average Height

-   [WIEGHT](#wieght) - Average Weight

-   [V00WTMAXKG](#v00wtmaxkg) - Self-reported max weight (kg)

-   [V00WTMINKG](#v00wtminkg) - Self-reported min weight (kg)

-   [Current Employment](#employment) (3 Columns)

    -   CEMPLOY_NWOR - Not working due to other
    -   CEMPLOY_NWH - Not working due to health
    -   CEMPLOY_FB - Unpaid work for family business
    -   CEMPLOY_W - Works for pay (reference)
    - This was simplified into working or not working

-   [Education Level](#education) (4 factor columns)

    -   V00EDCV_GradDeg - Graduate Degree
    -   V00EDCV_SomeGrad - Some Grad School
    -   V00EDCV_UGDeg - Undergrad Degree
    -   V00EDCV_SomeUG - Some Undergrad
    -   V00EDCV_HSDeg - High School Diploma
    -   V00EDCV_LessHS - No High School (reference)
    - This was simplified to remove someGrad/someUG

-   [NSAID](#nsaid) (factor) - Use of NSAIDs

-   [NARC](#narc) (factor) - Use of Narcotics

-   [Baseline X-ray Composite OA Grade](#p01OGRD) (4 factor columns)

    -   P01OAGRD_Severe
    -   P01OAGRD_Moderate
    -   P01OAGRD_Mild
    -   P01OAGRD_Possible
    -   P01OAGRD_Normal (reference)

-   [Hand Nodes](#hand-nodes) (2 Factor Columns)

    -   P02JBMPCV_NEW_None - no new hand nodules
    -   P02JBMPCV_NEW_One - New hand nodules on one hand
    -   P02JBMPCV_NEW_Both - New hand nodules on both hands (reference)

-   [COMORBSCORE](#comorbscore) - Charlson Score

-   [CESD](#cesd) - Depression Scale Score

    -   Converted to a binary for DPRESD (CESD \>= 16)

-   [SURG_INJ_HIST](#surg_inj_hist) (factor) - History of surgery or injury

The outcomes are as followed:

-   1: No event, no death

-   2: Left study before event

-   3: Death

-   4...8: Five-Level Random Forest Clusters

# Analysis of Predictors

### Number of Events

-   `r nrow(bsln[which(bsln$EVNT == 1),])` subjects with no event of interest
-   `r nrow(bsln[which(bsln$EVNT == 2),])` subjects who dropped out
-   `r nrow(bsln[which(bsln$EVNT == 3),])` subjects who died during the study
-   `r nrow(bsln[which(as.numeric(bsln$EVNT) >= 4),])` subjects who received a knee replacement
    -   `r nrow(bsln[which(bsln$EVNT == 4),])` in cluster 1
    -   `r nrow(bsln[which(bsln$EVNT == 5),])` in cluster 2
    -   `r nrow(bsln[which(bsln$EVNT == 6),])` in cluster 3
    -   `r nrow(bsln[which(bsln$EVNT == 7),])` in cluster 4
    -   `r nrow(bsln[which(bsln$EVNT == 8),])` in cluster 5

#### Age {#age}

```{r}
hist(bsln$AGE)
```


```{r}
bsln %>% 
    group_by(EVNT) %>%
    summarise(avg = mean(AGE), sd = sd(AGE), min = min(AGE), max = max(AGE)) %>%
    add_row(EVNT = "0", avg = mean(bsln$AGE), sd = sd(bsln$AGE), min = min(bsln$AGE), max = max(bsln$AGE)) %>%
    mutate(across(where(is.numeric), round, 2))
```

#### Sex {#sex}

```{r}
ggplot(bsln, aes(SEX, fill = EVNT)) + geom_bar()
```


```{r}
bsln %>% 
    group_by(EVNT) %>%
    summarize(freq = n(), male = sum(SEX == 1),
              male_perc = male/freq*100, 
              female = sum(SEX == 2), female_perc = female/freq*100) %>%
    add_row(EVNT = "0", freq = nrow(bsln), 
            male = sum(bsln$SEX == 1),  male_perc = male/freq*100, 
            female = sum(bsln$SEX == 2), female_perc = female/freq*100) %>%
    mutate(across(where(is.numeric), round, 2))
```

#### Race {#race}

```{r}
ggplot(bsln, aes(RACE_O, fill = EVNT)) + geom_bar()
```


```{r}
bsln %>% 
    group_by(EVNT) %>%
    summarize(freq = n(), NW = sum(RACE_O == 1), NW_perc = NW/freq*100, 
              W = freq - NW, W_perc = W / freq*100) %>%
    add_row(EVNT = "0", freq = nrow(bsln), 
            NW = sum(bsln$RACE_O == 1), NW_perc = NW/freq*100,
            W = freq - NW, W_perc = W / freq*100) %>%
    mutate(across(where(is.numeric), round, 2))
```

#### ETHNICITY {#ethnicity}

```{r}
ggplot(bsln, aes(ETHNICITY, fill = EVNT)) + geom_bar()
```

```{r}
bsln %>% 
    group_by(EVNT) %>%
    summarize(freq = n(), Hisp = sum(ETHNICITY == 1, na.rm = T), Hisp_perc = Hisp/freq*100) %>%
    add_row(EVNT = "0", freq = nrow(bsln), 
            Hisp = sum(bsln$ETHNICITY == 1, na.rm = T),  Hisp_perc = Hisp/freq*100) %>%
    mutate(across(where(is.numeric), round, 2))
```

#### PASE {#pase}

```{r}
par(mfrow = c(1, 2))
hist(bsln$PASE)
hist(log(bsln$PASE))
par(mfrow = c(1,1))
```


```{r echo=FALSE}
bsln %>% 
    filter(!is.na(PASE)) %>%
    group_by(EVNT) %>%
    summarise(avg = mean(PASE), sd = sd(PASE), min = min(PASE), max = max(PASE)) %>%
    add_row(EVNT = "0", avg = mean(bsln$PASE, na.rm = T), sd = sd(bsln$PASE, na.rm = T),
            min = min(bsln$PASE, na.rm = T), max = max(bsln$PASE, na.rm = T)) %>%
    mutate(across(where(is.numeric), round, 2))
```

#### MEDINS {#medins}

```{r}
ggplot(bsln, aes(MEDINS, fill = EVNT)) + geom_bar()
```

```{r}
bsln %>% 
    group_by(EVNT) %>%
    summarize(freq = n(), INS = sum(MEDINS == 1, na.rm = T), INS_perc = INS/freq*100) %>%
    add_row(EVNT = "0", freq = nrow(bsln), 
            INS = sum(bsln$MEDINS == 1, na.rm = T),  INS_perc = INS/freq*100) %>%
    mutate(across(where(is.numeric), round, 2))
```

#### WOMDL {#womdl}

```{r}
par(mfrow = c(1, 2))
hist(bsln$WOMADL)
hist(log(bsln$WOMADL + 1))
par(mfrow = c(1,1))
```

```{r}
bsln %>% 
    filter(!is.na(WOMADL)) %>%
    group_by(EVNT) %>%
    summarise(avg = mean(WOMADL), sd = sd(WOMADL), min = min(WOMADL), max = max(WOMADL)) %>%
    add_row(EVNT = "0", avg = mean(bsln$WOMADL, na.rm = T), sd = sd(bsln$WOMADL, na.rm = T), min = min(bsln$WOMADL, na.rm = T), max = max(bsln$WOMADL, na.rm = T))
```

#### WOMKP {#womkp}

```{r}
par(mfrow = c(1, 2))
hist(bsln$WOMKP)
hist(log(bsln$WOMKP + 1))
par(mfrow = c(1,1))
```

```{r}
bsln %>% 
    filter(!is.na(WOMKP)) %>%
    group_by(EVNT) %>%
    summarise(avg = mean(WOMKP), sd = sd(WOMKP), min = min(WOMKP), max = max(WOMKP)) %>%
    add_row(EVNT = "0", avg = mean(bsln$WOMKP, na.rm = T), sd = sd(bsln$WOMKP, na.rm = T), min = min(bsln$WOMKP, na.rm = T), max = max(bsln$WOMKP, na.rm = T))
```

#### WOMSTF {#womstf}

```{r}
par(mfrow = c(1, 2))
hist(bsln$WOMSTF)
hist(log(bsln$WOMSTF + 1))
par(mfrow = c(1,1))
```


```{r}
bsln %>% 
    filter(!is.na(WOMSTF)) %>%
    group_by(EVNT) %>%
    summarise(avg = mean(WOMSTF), sd = sd(WOMSTF), min = min(WOMSTF), max = max(WOMSTF)) %>%
    add_row(EVNT = "0", avg = mean(bsln$WOMSTF, na.rm = T), sd = sd(bsln$WOMSTF, na.rm = T), min = min(bsln$WOMSTF, na.rm = T), max = max(bsln$WOMSTF, na.rm = T))
```

#### HEIGHT

```{r}
hist(bsln$HEIGHT)
```

```{r}
bsln %>% 
    filter(!is.na(HEIGHT)) %>%
    group_by(EVNT) %>%
    summarise(avg = mean(HEIGHT), sd = sd(HEIGHT), min = min(HEIGHT), max = max(HEIGHT)) %>%
    add_row(EVNT = "0", avg = mean(bsln$HEIGHT, na.rm = T), sd = sd(bsln$HEIGHT, na.rm = T), min = min(bsln$HEIGHT, na.rm = T), max = max(bsln$HEIGHT, na.rm = T))
```

#### Weight

```{r}
hist(bsln$WEIGHT)
```

```{r}
bsln %>% 
    filter(!is.na(WEIGHT)) %>%
    group_by(EVNT) %>%
    summarise(avg = mean(WEIGHT), sd = sd(WEIGHT), min = min(WEIGHT), max = max(WEIGHT)) %>%
    add_row(EVNT = "0", avg = mean(bsln$WEIGHT, na.rm = T), sd = sd(bsln$WEIGHT, na.rm = T), min = min(bsln$WEIGHT, na.rm = T), max = max(bsln$WEIGHT, na.rm = T))
```

#### BMI {#bmi}

```{r}
hist(bsln$BMI)
```

```{r}
bsln %>% 
    filter(!is.na(BMI)) %>%
    group_by(EVNT) %>%
    summarise(avg = mean(BMI), sd = sd(BMI), min = min(BMI), max = max(BMI)) %>%
    add_row(EVNT = "0", avg = mean(bsln$BMI, na.rm = T), sd = sd(bsln$BMI, na.rm = T), min = min(bsln$BMI, na.rm = T), max = max(bsln$BMI, na.rm = T))
```

#### V00WTMAXKG {#v00wtmaxkg}

```{r}
hist(bsln$V00WTMAXKG)
```

```{r}
bsln %>% 
    filter(!is.na(V00WTMAXKG)) %>%
    group_by(EVNT) %>%
    summarise(avg = mean(V00WTMAXKG), sd = sd(V00WTMAXKG), min = min(V00WTMAXKG), max = max(V00WTMAXKG)) %>%
    add_row(EVNT = "0", avg = mean(bsln$V00WTMAXKG, na.rm = T), sd = sd(bsln$V00WTMAXKG, na.rm = T), min = min(bsln$V00WTMAXKG, na.rm = T), max = max(bsln$V00WTMAXKG, na.rm = T))
```

#### V00WTMIXKG {#v00wtmixkg}

```{r}
par(mfrow = c(1, 2))
hist(bsln$V00WTMINKG)
hist(log(bsln$V00WTMINKG))
par(mfrow = c(1,1))
```

```{r}
bsln %>% 
    filter(!is.na(V00WTMINKG)) %>%
    group_by(EVNT) %>%
    summarise(avg = mean(V00WTMINKG), sd = sd(V00WTMINKG), min = min(V00WTMINKG), max = max(V00WTMINKG)) %>%
    add_row(EVNT = "0", avg = mean(bsln$V00WTMINKG, na.rm = T), sd = sd(bsln$V00WTMINKG, na.rm = T), min = min(bsln$V00WTMINKG, na.rm = T), max = max(bsln$V00WTMINKG, na.rm = T))
```

#### Employment {#employment}

```{r}
ggplot(bsln, aes(CEMPLOY_NW, fill = EVNT)) + geom_bar()
```

```{r}
bsln %>% 
    group_by(EVNT) %>%
    summarize(freq = n(), NOWORK = sum(CEMPLOY_NW == 1), NW_perc = NOWORK/freq*100) %>%
    add_row(EVNT = "0", freq = nrow(bsln), 
            NOWORK = sum(bsln$CEMPLOY_NW == 1),  NW_perc = NOWORK/freq*100) %>%
    mutate(across(where(is.numeric), round, 2))
```

#### Education {#education}

```{r}
ggplot(bsln, aes(EDCV, fill=EVNT)) + geom_bar()
```

```{r}
bsln %>% 
  group_by(EVNT) %>%
  summarise(freq = n(), 
            Grad = sum(EDCV_GradDeg == 1), grd_per = Grad / freq*100,
            Undergrad = sum(EDCV_UGDeg == 1), ug_per = Undergrad / freq*100,
            High_school = sum(EDCV_HSDeg == 1), pos_per = High_school / freq*100,
            None = freq - Grad - Undergrad - High_school, non_per = None / freq*100
            ) %>%
    add_row(EVNT = "0", freq = nrow(bsln), 
            Grad = sum(bsln$EDCV_GradDeg == 1), grd_per = Grad / freq*100,
            Undergrad = sum(bsln$EDCV_UGDeg == 1), ug_per = Undergrad / freq*100,
            High_school = sum(bsln$EDCV_HSDeg == 1), pos_per = High_school / freq*100,
            None = freq - Grad - Undergrad - High_school, non_per = None / freq*100) %>%
    mutate(across(where(is.numeric), round, 2))
```

#### NSAID {#nsaid}

```{r}
ggplot(bsln, aes(NSAID, fill = EVNT)) + geom_bar()
```

```{r}
bsln %>% 
  filter(!is.na(NSAID)) %>%
  group_by(EVNT) %>%
  summarise(freq = n(), NSAID = sum(NSAID == 1), per = NSAID / freq*100) %>%
  add_row(EVNT = "0", freq = nrow(bsln %>% filter(!is.na(NSAID))), 
            NSAID = sum(bsln$NSAID == 1, na.rm = T),  per = NSAID/freq*100) %>% 
    mutate(across(where(is.numeric), round, 2))
```

#### NARC {#narc}

```{r}
ggplot(bsln, aes(NARC, fill = EVNT)) + geom_bar()
```

```{r}
bsln %>% 
  filter(!is.na(NARC)) %>%
  group_by(EVNT) %>%
  summarise(freq = n(), NARC = sum(NARC == 1), per = NARC / freq*100) %>%
  add_row(EVNT = "0", freq = nrow(bsln %>% filter(!is.na(NARC))), 
            NARC = sum(bsln$NARC == 1, na.rm = T),  per = NARC/freq*100) %>% 
    mutate(across(where(is.numeric), round, 2))
```

#### P01OGRD {#p01ogrd}

```{r}
ggplot(bsln, aes(P01OAGRD, fill = EVNT)) + geom_bar()
```

```{r}
bsln %>% 
  group_by(EVNT) %>%
  summarise(freq = n(), 
            Severe = sum(P01OAGRD_Severe == 1), sev_per = Severe / freq*100,
            Moderate = sum(P01OAGRD_Moderate == 1), mod_per = Moderate / freq*100,
            Possible = sum(P01OAGRD_Possible == 1), pos_per = Possible / freq*100,
            Mild = sum(P01OAGRD_Mild == 1), mld_per = Mild / freq*100,
            None = freq - Severe - Moderate - Possible - Mild, non_per = None / freq*100
            ) %>%
  add_row(EVNT = "0", freq = nrow(bsln), 
            Severe = sum(bsln$P01OAGRD_Severe == 1),  sev_per = Severe/freq*100, 
            Moderate = sum(bsln$P01OAGRD_Moderate == 1), mod_per = Moderate/freq*100,
            Possible = sum(bsln$P01OAGRD_Possible == 1), pos_per = Possible/freq*100,
            Mild = sum(bsln$P01OAGRD_Mild == 1), mld_per = Mild/freq*100,
            None = freq - Severe - Moderate - Possible - Mild, non_per = None / freq*100) %>%
    mutate(across(where(is.numeric), round, 2))
```

#### Hand Nodes

```{r}
ggplot(bsln, aes(P02JBMPCV_NEW, fill = EVNT)) + geom_bar()
```

```{r}
bsln %>% 
  group_by(EVNT) %>%
  summarise(freq = n(), 
            None = sum(P02JBMPCV_NEW_None == 1), None_per = None / freq*100,
            One = sum(P02JBMPCV_NEW_One == 1), One_per = One / freq*100,
            Both = freq - One - None, Both_per = Both / freq*100
            ) %>%
  add_row(EVNT = "0", freq = nrow(bsln), 
            None = sum(bsln$P02JBMPCV_NEW_None == 1),  None_per = None/freq*100, 
            One = sum(bsln$P02JBMPCV_NEW_One == 1), One_per = One/freq*100,
            Both = freq - One - None, Both_per = Both / freq*100) %>%
    mutate(across(where(is.numeric), round, 2))
```

#### COMORBSCORE

```{r}
par(mfrow = c(1, 2))
hist(bsln$COMORBSCORE)
hist(log(bsln$COMORBSCORE))
par(mfrow = c(1,1))
```

```{r}
bsln %>% 
    filter(!is.na(COMORBSCORE)) %>%
    group_by(EVNT) %>%
    summarise(avg = mean(COMORBSCORE), sd = sd(COMORBSCORE), min = min(COMORBSCORE), max = max(COMORBSCORE)) %>%
    add_row(EVNT = "0", avg = mean(bsln$COMORBSCORE, na.rm = T), sd = sd(bsln$COMORBSCORE, na.rm = T), min = min(bsln$COMORBSCORE, na.rm = T), max = max(bsln$COMORBSCORE, na.rm = T)) %>%
    mutate(across(where(is.numeric), round, 2))
```

#### CESD

```{r}
par(mfrow = c(1, 2))
hist(bsln$CESD)
hist(log(bsln$CESD + 1))
par(mfrow = c(1,1))
```

```{r}
bsln %>% 
    filter(!is.na(CESD)) %>%
    group_by(EVNT) %>%
    summarise(avg = mean(CESD), sd = sd(CESD), min = min(CESD), max = max(CESD)) %>%
    add_row(EVNT = "0", avg = mean(bsln$CESD, na.rm = T), sd = sd(bsln$CESD, na.rm = T), min = min(bsln$CESD, na.rm = T), max = max(bsln$CESD, na.rm = T)) %>%
    mutate(across(where(is.numeric), round, 2))
```

#### DPRSD

```{r}
ggplot(bsln, aes(DPRSD, fill = EVNT)) + geom_bar()
```

```{r}
bsln %>% 
    group_by(EVNT) %>%
    summarize(freq = n(), Dprsd = sum(DPRSD == 1, na.rm = T), Dprsd_perc = Dprsd/freq*100) %>%
    add_row(EVNT = "0", freq = nrow(bsln), 
            Dprsd = sum(bsln$DPRSD == 1, na.rm = T),  Dprsd_perc = Dprsd/freq*100) %>%
    mutate(across(where(is.numeric), round, 2))
```

#### Surg_Inj_Hist {#surg_inj_hist}

```{r}
ggplot(bsln, aes(Surg_Inj_Hist, fill = EVNT)) + geom_bar()
```

```{r}
bsln %>% 
  filter(!is.na(Surg_Inj_Hist)) %>% 
  group_by(EVNT) %>%
  summarise(freq = n(), Surg_Inj_Hist = sum(Surg_Inj_Hist == 1), per = Surg_Inj_Hist / freq*100) %>%
  add_row(EVNT = "0", freq = nrow(bsln %>% filter(!is.na(Surg_Inj_Hist))), 
            Surg_Inj_Hist = sum(bsln$Surg_Inj_Hist == 1, na.rm = T),  per = Surg_Inj_Hist/freq*100) %>% 
    mutate(across(where(is.numeric), round, 2))
```

## Corelation Between Variables

```{r}
library(corrplot)
library(RColorBrewer)
num_col <-  c("AGE", "PASE", "WOMADL", "WOMKP", "WOMSTF", "V00WTMAXKG", 
                 "V00WTMINKG", "BMI", "HEIGHT", "WEIGHT", "COMORBSCORE", "CESD")

M <-cor(bsln[,num_col], use = "pairwise.complete.obs")
corrplot(M, type="upper", order="hclust",
         col=brewer.pal(n=8, name="RdYlBu"))
```