There are 24 variables of interest in this data set:

-   ID

-   [AGE](#age)

-   [SEX](#sex) (1 = Male, 2 = Female)

-   [Race](#race) (2 factor columns)

    -   RACE_AA - African-American
    -   RACE_NW - Other Non-White
    -   White (reference)

-   [ETHNICITY](#ethnicity) (factor) - Reported Hispanic or latino

-   [PASE](#pase) - Physical Activity Score for the Elderly

-   [MEDINS](#medins) - Insurance prescript payment (factor)

-   [WOMADL](#womadl) - WOMAC Disability Score

-   [WOMKP](#womkp) - WOMAC Pain Score

-   [WOMSTF](#womstf) - WOMAAC Stiffness Score

-   [BMI](#bmi)

-   [HEIGHT](#height) - Average Height

-   [WIEGHT](#wieght) - Average Weight

-   [V00WTMAXKG](#v00wtmaxkg) - Self-reported max weight (kg)

-   [V00WTMINKG](#v00wtminkg) - Self-reported min weight (kg)

-   [Current Employment](#employment) (3 Columns)

    -   CEMPLOY_NWOR - Not working due to other
    -   CEMPLOY_NWH - Not working due to health
    -   CEMPLOY_FB - Unpaid work for family business
    -   CEMPLOY_W - Works for pay (reference)

-   [Education Level](#education) (4 factor columns)

    -   V00EDCV_GradDeg - Graduate Degree
    -   V00EDCV_SomeGrad - Some Grad School
    -   V00EDCV_UGDeg - Undergrad Degree
    -   V00EDCV_SomeUG - Some Undergrad
    -   V00EDCV_HSDeg - High School Diploma
    -   V00EDCV_LessHS - No High School (reference)

-   [NSAID](#nsaid) (factor) - Use of NSAIDs

-   [NARC](#narc) (factor) - Use of Narcotics

-   [Baseline X-ray Composite OA Grade](#p01OGRD) (4 factor columns)

    -   P01OAGRD_Severe
    -   P01OAGRD_Moderate
    -   P01OAGRD_Mild
    -   P01OAGRD_Possible
    -   P01OAGRD_Normal (reference)

-   [Hand Nodes](#hand-nodes) (2 Factor Columns)

    -   P02JBMPCV_NEW_None - no new hand nodules
    -   P02JBMPCV_NEW_One - New hand nodules on one hand
    -   P02JBMPCV_NEW_Both - New hand nodules on both hands (reference)

-   [COMORBSCORE](#comorbscore) - Charlson Score

-   [CESD](#cesd) - Depression Scale Score

    -   Converted to a binary for DPRESD (CESD \>= 16)

-   [SURG_INJ_HIST](#surg_inj_hist) (factor) - History of surgery or injury

The outcomes are as followed:

-   1: No event, no death

-   2: Left study before event

-   3: Death

-   4...8: Five-Level Random Forest Clusters

# Analysis of Predictors

### Number of Events

-   `r nrow(data_full[which(data_full$EVNT == 1),])` subjects with no event of interest
-   `r nrow(data_full[which(data_full$EVNT == 2),])` subjects who dropped out
-   `r nrow(data_full[which(data_full$EVNT == 3),])` subjects who died during the study
-   `r nrow(data_full[which(data_full$EVNT >= 4),])` subjects who received a knee replacement
    -   `r nrow(data_full[which(data_full$EVNT == 4),])` in cluster 1
    -   `r nrow(data_full[which(data_full$EVNT == 5),])` in cluster 2
    -   `r nrow(data_full[which(data_full$EVNT == 6),])` in cluster 3
    -   `r nrow(data_full[which(data_full$EVNT == 7),])` in cluster 4
    -   `r nrow(data_full[which(data_full$EVNT == 8),])` in cluster 5

#### Age {#age}

```{r}
data_full %>% 
    group_by(EVNT) %>%
    summarise(avg = mean(AGE), sd = sd(AGE), min = min(AGE), max = max(AGE)) %>%
    add_row(EVNT = 0, avg = mean(data_full$AGE), sd = sd(data_full$AGE), min = min(data_full$AGE), max = max(data_full$AGE))
```

#### Sex {#sex}

```{r}
data_full %>% 
    group_by(EVNT) %>%
    summarize(freq = n(), male = sum(SEX == 1),
              male_perc = male/freq*100, 
              female = sum(SEX == 2), female_perc = female/freq*100) %>%
    add_row(EVNT = 0, freq = nrow(data_full), 
            male = sum(data_full$SEX == 1),  male_perc = male/freq*100, 
            female = sum(data_full$SEX == 2), female_perc = female/freq*100) %>%
    mutate(across(where(is.numeric), round, 2))

```

#### Race {#race}

```{r}
data_full %>% 
    group_by(EVNT) %>%
    summarize(freq = n(), AA = sum(RACE_AA == 1),AA_perc = AA/freq*100, 
              NW = sum(RACE_NW == 1), NW_perc = NW/freq*100, 
              W = freq - AA - NW, W_perc = W / freq*100) %>%
    add_row(EVNT = 0, freq = nrow(data_full), 
            AA = sum(data_full$RACE_AA == 1),  AA_perc = AA/freq*100, 
            NW = sum(data_full$RACE_NW == 1), NW_perc = NW/freq*100,
            W = freq - AA - NW, W_perc = W / freq*100) %>%
    mutate(across(where(is.numeric), round, 2))
```

#### PASE {#pase}

```{r}
data_full %>% 
    group_by(EVNT) %>%
    summarise(avg = mean(PASE), sd = sd(PASE), min = min(PASE), max = max(PASE)) %>%
    add_row(EVNT = 0, avg = mean(data_full$PASE), sd = sd(data_full$PASE), min = min(data_full$PASE), max = max(data_full$PASE))
```

#### MEDINS {#medins}

```{r}
data_full %>% 
    group_by(EVNT) %>%
    summarise(avg = mean(MEDINS), sd = sd(MEDINS), min = min(MEDINS), max = max(MEDINS)) %>%
    add_row(EVNT = 0, avg = mean(data_full$MEDINS), sd = sd(data_full$MEDINS), min = min(data_full$MEDINS), max = max(data_full$MEDINS))
```

#### WOMKP {#womkp}

```{r}
data_full %>% 
    group_by(EVNT) %>%
    summarise(avg = mean(WOMKP), sd = sd(WOMKP), min = min(WOMKP), max = max(WOMKP)) %>%
    add_row(EVNT = 0, avg = mean(data_full$WOMKP), sd = sd(data_full$WOMKP), min = min(data_full$WOMKP), max = max(data_full$WOMKP))
```

#### WOMSTF {#womstf}

```{r}
data_full %>% 
    group_by(EVNT) %>%
    summarise(avg = mean(WOMSTF), sd = sd(WOMSTF), min = min(WOMSTF), max = max(WOMSTF)) %>%
    add_row(EVNT = 0, avg = mean(data_full$WOMSTF), sd = sd(data_full$WOMSTF), min = min(data_full$WOMSTF), max = max(data_full$WOMSTF))
```

#### Weight

```{r}
data_full %>% 
    group_by(EVNT) %>%
    summarise(avg = mean(WEIGHT), sd = sd(WEIGHT), min = min(WEIGHT), max = max(WEIGHT)) %>%
    add_row(EVNT = 0, avg = mean(data_full$WEIGHT), sd = sd(data_full$WEIGHT), min = min(data_full$WEIGHT), max = max(data_full$WEIGHT))
```

#### BMI {#bmi}

```{r}
data_full %>% 
    group_by(EVNT) %>%
    summarise(avg = mean(BMI), sd = sd(BMI), min = min(BMI), max = max(BMI)) %>%
    add_row(EVNT = 0, avg = mean(data_full$BMI), sd = sd(data_full$BMI), min = min(data_full$BMI), max = max(data_full$BMI))
```

#### V00WTMAXKG {#v00wtmaxkg}

```{r}
data_full %>% 
    group_by(EVNT) %>%
    summarise(avg = mean(V00WTMAXKG), sd = sd(V00WTMAXKG), min = min(V00WTMAXKG), max = max(V00WTMAXKG)) %>%
    add_row(EVNT = 0, avg = mean(data_full$V00WTMAXKG), sd = sd(data_full$V00WTMAXKG), min = min(data_full$V00WTMAXKG), max = max(data_full$V00WTMAXKG))
```

#### Employment {#employment}

```{r}
data_full %>% 
    group_by(EVNT) %>%
    summarize(freq = n(), NWOR = sum(CEMPLOY_NWOR == 1), NWOR_perc = NWOR/freq*100, 
              Other = freq - NWOR, Other_perc = Other / freq*100) %>%
    add_row(EVNT = 0, freq = nrow(data_full), 
            NWOR = sum(data_full$CEMPLOY_NWOR == 1),  NWOR_perc = NWOR/freq*100, 
            Other = freq - NWOR, Other_perc = Other / freq*100) %>%
    mutate(across(where(is.numeric), round, 2))
```

#### Education {#education}

```{r}
data_full %>% 
  group_by(EVNT) %>%
  summarise(freq = n(), 
            Grad = sum(EDCV_GradDeg == 1), grd_per = Grad / freq*100,
            Undergrad = sum(EDCV_UGDeg == 1), ug_per = Undergrad / freq*100,
            High_school = sum(EDCV_HSDeg == 1), pos_per = High_school / freq*100,
            None = freq - Grad - Undergrad - High_school, non_per = None / freq*100
            ) %>%
    add_row(EVNT = 0, freq = nrow(data_full), 
            Grad = sum(data_full$EDCV_GradDeg == 1), grd_per = Grad / freq*100,
            Undergrad = sum(data_full$EDCV_UGDeg == 1), ug_per = Undergrad / freq*100,
            High_school = sum(data_full$EDCV_HSDeg == 1), pos_per = High_school / freq*100,
            None = freq - Grad - Undergrad - High_school, non_per = None / freq*100) %>%
    mutate(across(where(is.numeric), round, 2))
```

#### NSAID {#nsaid}

```{r}
data_full %>% 
  group_by(EVNT) %>%
  summarise(freq = n(), NSAID = sum(NSAID == 1), per = NSAID / freq*100) %>%
  add_row(EVNT = 0, freq = nrow(data_full), 
            NSAID = sum(data_full$NSAID == 1),  per = NSAID/freq*100) %>% 
    mutate(across(where(is.numeric), round, 2))
```

#### P01OGRD {#p01ogrd}

```{r}
data_full %>% 
  group_by(EVNT) %>%
  summarise(freq = n(), 
            Severe = sum(P01OAGRD_Severe == 1), sev_per = Severe / freq*100,
            Moderate = sum(P01OAGRD_Moderate == 1), mod_per = Moderate / freq*100,
            Possible = sum(P01OAGRD_Possible == 1), pos_per = Possible / freq*100,
            Mild = sum(P01OAGRD_Mild == 1), mld_per = Mild / freq*100,
            None = freq - Severe - Moderate - Possible - Mild, non_per = None / freq*100
            ) %>%
  add_row(EVNT = 0, freq = nrow(data_full), 
            Severe = sum(data_full$P01OAGRD_Severe == 1),  sev_per = Severe/freq*100, 
            Moderate = sum(data_full$P01OAGRD_Moderate == 1), mod_per = Moderate/freq*100,
            Possible = sum(data_full$P01OAGRD_Possible == 1), pos_per = Possible/freq*100,
            Mild = sum(data_full$P01OAGRD_Mild == 1), mld_per = Mild/freq*100,
            None = freq - Severe - Moderate - Possible - Mild, non_per = None / freq*100) %>%
    mutate(across(where(is.numeric), round, 2))
```

#### Surg_Inj_Hist {#surg_inj_hist}

```{r}
data_full %>% 
  group_by(EVNT) %>%
  summarise(freq = n(), Surg_Inj_Hist = sum(Surg_Inj_Hist == 1), per = Surg_Inj_Hist / freq*100) %>%
  add_row(EVNT = 0, freq = nrow(data_full), 
            Surg_Inj_Hist = sum(data_full$Surg_Inj_Hist == 1),  per = Surg_Inj_Hist/freq*100) %>% 
    mutate(across(where(is.numeric), round, 2))
```
