---
title: "OAI - Model Selection - K5 with BMI"
author: "Mitchell Henschel"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: true
    theme: journal
---

## Settings

```{r settings, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(tidyverse)
library(nnet) # For multinomial model
options(scipen = 999) # No scientific notation
```

## Data Loading

```{r loadData, message=FALSE, warning=FALSE}
DATAPATH <- Sys.getenv("OAI_DATA")
if (DATAPATH == "" ) stop( "Please add datapath to OAI_DATA" )

source("OAI_LoadData.R", chdir = T) # LoadData file must be in working directory

if(!exists('data_full')) {
  
  if(!exists('bsln')) {
    bsln <- getBaselineData(DATAPATH)
  }
  
  if(!exists('evnt')) {
    evnt <- getEvents(DATAPATH)
  }
  
  data_k5 <- getCompleteData(DATAPATH, "K.5.Clusters")
  data_full <- data_k5
  remove(data_k5)
  remove(bsln)
  remove(evnt)
}
```

#### Missing Data:

-   There are `r sum(!complete.cases(data_full))` entries in with missing data.
-   `r sum(!complete.cases(data_full[as.numeric(data_full$EVNT) >= 4,]))` who received knee replacements
-   `r sum(!complete.cases(data_full[as.numeric(data_full$EVNT) < 4,]))` who did not

These observations will be removed from the data

```{r omitNa}
data_full <- na.omit(data_full)
```

# Model Selection

After trial and error we removed height, min weight, comorbity score, and binary depression as they had little value in prediction.

```{r fullMod, results=FALSE}
predictors_full <- c("AGE", "SEX", "MEDINS", "PASE", "WOMADL", "WOMKP", "WOMSTF", "V00WTMAXKG", "WEIGHT", "HEIGHT", "CESD", "NSAID", "NARC", "Surg_Inj_Hist", "CEMPLOY_NW", "EDCV_GradDeg", "EDCV_UGDeg", "EDCV_HSDeg", "P01OAGRD_Severe", "P01OAGRD_Moderate", "P01OAGRD_Mild", "P01OAGRD_Possible", "P02JBMPCV_NEW_None", "P02JBMPCV_NEW_One", "RACE_O")
eqtn_full <- formula(paste("EVNT ~ ", paste(predictors_full, collapse = " + "), " + WEIGHT:HEIGHT"))
mod_full <- multinom(eqtn_full, data=data_full, maxit = 1000)
dev_full <- deviance(mod_full)
df_full <- mod_full$edf
aic_full <- mod_full$AIC
```

Residual Deviance: `r dev_full`

AIC: `r aic_full`

It was decided that the base model should always contain age, sex, WOMAC pain score, and BMI. The multinom function from the nnet package is used.

```{r baseMod, results=FALSE}
mod_base <- multinom(EVNT ~ AGE + SEX + WOMKP + WEIGHT + HEIGHT + WEIGHT:HEIGHT, data=data_full)
dev_base <- deviance(mod_base)
df_base <- mod_base$edf
aic_base <- mod_base$AIC
p_full <- pchisq(dev_base - dev_full, df_full - df_base,lower = F)
```

Residual Deviance: `r df_base`

AIC: `r aic_base`

P-Val of Deviance Diff: `r p_full`

### Stepwise Selection

```{r forward, results=FALSE}
step_forward <- step(mod_base, direction = 'forward', scope=formula(mod_full), trace=0, scale = 'glm')
dev_for <- deviance(step_forward)
df_for <- step_forward$edf
aic_for <- step_forward$AIC
p_for <- pchisq(dev_for - dev_full, df_full - df_for, lower.tail = F)
```

-   Forward Selection

    Residual Deviance: `r dev_for`
    
    AIC: `r aic_for`
    
    P-Val of Deviance Diff: `r p_for`
    
    Selected Predictors:
    `r step_forward$anova$Step`

```{r backward, results=FALSE}
step_backward <- step(mod_full, direction = 'backward', scope=formula(mod_full), trace=0, scale = 'glm')
dev_bac <- deviance(step_backward)
df_bac <- step_backward$edf
aic_bac <- step_backward$AIC
p_bac <- pchisq(dev_bac-dev_full, df_full - df_bac, lower.tail = F)
```

-   Backward Selection

    Residual Deviance: `r dev_bac`
    
    AIC: `r aic_bac`
    
    P-Val of Deviance Diff: `r p_bac`
    
    Selected Predictors:
    `r step_backward$anova$Step`

```{r both, results=FALSE}
beststep_both <- step(mod_base, direction = 'both', scope=formula(mod_full), trace=0, scale = 'glm')
dev_both <- deviance(beststep_both)
df_both <- beststep_both$edf
aic_both <- beststep_both$AIC
p_both <- pchisq(dev_both-dev_full, df_full - df_both, lower.tail = F)
```

-   Both direction selection

    Residual Deviance: `r dev_both`
    
    AIC: `r aic_both`
    
    P-Val of Deviance Diff: `r p_both`
    
    Selected Predictors:
    `r beststep_both$anova$Step`