---
title: 'BS852 Class 10'
author: ""
date: ""
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# set working directory if necessary:
knitr::opts_knit$set(root.dir = "C:\\")

library(effects) # To plot interactions
library(survival) # load survival package

```

## FHS Example

#### This chunk of code 
* reads data, 
* extract records with complete data
* selects records for *FEMALES* and those with *CHD=0 OR CHD>4*
```{r }
   
framdat2 <- read.table("framdat2.txt",header=T, na.strings=c("."))

work.data <- na.omit(framdat2[,-c(3,4)]) ## drop DTH and CAU columns and remove rows with missing data 
work.data <- subset(work.data, (work.data$SEX == 2) & (work.data$CHD==0 | work.data$CHD > 4))
work.data$chd_sw = work.data$CHD >= 4  ## 1 is event, 0 is no event
``` 


#### Model with GLI and AGE
```{r }
work.data$GLI <- as.factor(work.data$GLI)
mod.age.adjusted <- glm(chd_sw ~ GLI + AGE, family=binomial, data=work.data)
summary(mod.age.adjusted)
plot(allEffects(mod.age.adjusted))

```


#### Interpretation of fitted values
```{r }
log.odds.CHD <- log(mod.age.adjusted$fitted.values/(1-mod.age.adjusted$fitted.values))
plot(work.data$AGE,log.odds.CHD,xlab="Age", ylab="log-odds-CHD")
points(work.data$AGE[work.data$GLI==1],log.odds.CHD[work.data$GLI==1],col=2)

odds.CHD <- mod.age.adjusted$fitted.values/(1-mod.age.adjusted$fitted.values)
plot(work.data$AGE,odds.CHD,xlab="Age", ylab="odds-CHD")
points(work.data$AGE[work.data$GLI==1],odds.CHD[work.data$GLI==1],col=2)
   
```

####  Test Interaction Between Age and GLI
```{r }

mod.interaction <- glm(chd_sw ~ GLI*AGE, family = binomial, data=work.data)
summary(mod.interaction)
plot(allEffects(mod.interaction))
plot(allEffects(mod.interaction), multiline=TRUE)

log.odds.CHD <- log( mod.interaction$fitted.values/(1-mod.interaction$fitted.values))   
plot(work.data$AGE,log.odds.CHD,xlab="Age", ylab="log-odds-CHD")
points(work.data$AGE[work.data$GLI==1],log.odds.CHD[work.data$GLI==1],col=2)

odds.CHD <- mod.interaction$fitted.values/(1-mod.interaction$fitted.values)
plot(work.data$AGE,odds.CHD,xlab="Age", ylab="odds-CHD")
points(work.data$AGE[work.data$GLI==1],odds.CHD[work.data$GLI==1],col=2)

```

## Cox PH regression example
```{r}

cvd <- read.csv("CVD.csv", header=T, na.strings=c("."))
cvd$studytime <- cvd$Age.last.contact - cvd$Age.enrollment
cvd$studytime[which(cvd$CVD==1)] <- cvd$CVD.inc.age[which(cvd$CVD==1)] - cvd$Age.enrollment[which(cvd$CVD==1)]

```

```{r}

mod.ph <- coxph(Surv(studytime, CVD) ~ DSST*sex + bmi + sysbp + Age.enrollment, data=cvd)
summary(mod.ph)

# aHR for 1-unit change in DSST (among males):
exp(c(HR = coef(mod.ph)[1], confint.default(mod.ph)[1,])) 
# HRR.DSST     2.5 %    97.5 % 
# 1.0022175 0.9922614 1.0122735

# aHR for 1-unit change in DSST (among females):
exp(c(HR=(coef(mod.ph)[1]+coef(mod.ph)[6]), "2.5%"=(coef(mod.ph)[1]+coef(mod.ph)[6])-1.96*sqrt(vcov(mod.ph)[1,1]+vcov(mod.ph)[6,6]+2*vcov(mod.ph)[1,6]),
          "97.5%"=(coef(mod.ph)[1]+coef(mod.ph)[6])+1.96*sqrt(vcov(mod.ph)[1,1]+vcov(mod.ph)[6,6]+2*vcov(mod.ph)[1,6])))
#   HR.DSST  2.5%.DSST 97.5%.DSST 
#  1.015680   1.004880   1.026596 


# aHR for 10-unit change in DSST (among males):
exp(10*c(HR = coef(mod.ph)[1], confint.default(mod.ph)[1,])) 
#   HR.DSST     2.5 %    97.5 % 
#  1.0223979 0.9252542 1.1297408

# aHR for 10-unit change in DSST (among females):
exp(10*c(HR=(coef(mod.ph)[1]+coef(mod.ph)[6]), "2.5%"=(coef(mod.ph)[1]+coef(mod.ph)[6])-1.96*sqrt(vcov(mod.ph)[1,1]+vcov(mod.ph)[6,6]+2*vcov(mod.ph)[1,6]),
          "97.5%"=(coef(mod.ph)[1]+coef(mod.ph)[6])+1.96*sqrt(vcov(mod.ph)[1,1]+vcov(mod.ph)[6,6]+2*vcov(mod.ph)[1,6])))
#  HR.DSST  2.5%.DSST 97.5%.DSST 
#  1.168339   1.049884   1.300159 

```


## Interactions with two covariates

### Example #1

#### Create data
```{r}

data <- rbind( 
  cbind( rep(0, 100), rep(0,  100), rep(0, 100), rep(0, 100)),
  cbind( rep(0, 200), rep(1,  200), rep(0, 200), rep(0, 200)),
  cbind( rep(1,  40), rep(0,   40), rep(0,  40), rep(0,  40)),
  cbind( rep(1,1280), rep(1, 1280), rep(0,1280), rep(0,1280)),
  cbind( rep(0, 200), rep(0,  200), rep(1, 200), rep(0, 200)),
  cbind( rep(0, 100), rep(1,  100), rep(1, 100), rep(0, 100)),
  cbind( rep(1,  50), rep(0,   50), rep(1,  50), rep(0,  50)),
  cbind( rep(1, 200), rep(1,  200), rep(1, 200), rep(0,  200)),
  cbind( rep(0,  80), rep(0,   80), rep(0,  80), rep(1,  80)),
  cbind( rep(0, 100), rep(1,  100), rep(0, 100), rep(1, 100)),
  cbind( rep(1,  20), rep(0,   20), rep(0,  20), rep(1,  20)),
  cbind( rep(1, 250), rep(1,  250), rep(0, 250), rep(1, 250)),
  cbind( rep(0, 160), rep(0,  160), rep(1, 160), rep(1, 160)),
  cbind( rep(0, 400), rep(1,  400), rep(1, 400), rep(1, 400)),
  cbind( rep(1,  50), rep(0,   50), rep(1,  50), rep(1,  50)),
  cbind( rep(1, 625), rep(1,  625), rep(1, 625), rep(1, 625)))
data <- as.data.frame(data)
names(data) <- c("End.Ca", "Estr", "HxGall", "Hyper") 
data[,names(data)] <- lapply(data[,names(data)] , factor)
table(data)

```

#### Full model
```{r}
mod.full <- glm(End.Ca ~ Estr*HxGall*Hyper,
                  data=data, family=binomial)
summary(mod.full)

```


#### Drop 3-way interaction, HxGall*Hyper
```{r}
mod.drop <- glm(End.Ca ~ Estr*HxGall +Estr*Hyper,
                 data=data, family=binomial)
summary(mod.drop)
plot(allEffects(mod.drop), multiline=TRUE)
```

#### ORs: reference group is women without ERT and no hypertension and no history of gall bladder disease
```{r}

OR100 <- exp(coef(mod.drop)[2])
OR110 <- exp(coef(mod.drop)[2]+coef(mod.drop)[3]+coef(mod.drop)[5])
OR101 <- exp(coef(mod.drop)[2]+coef(mod.drop)[4]+coef(mod.drop)[6])
OR111 <- exp(coef(mod.drop)[2]+coef(mod.drop)[3]+coef(mod.drop)[4]+
               coef(mod.drop)[5]+coef(mod.drop)[6])

OR100
OR110
OR101
OR111
```

## Interactions with two covariates

### Example #2

#### Create data
```{r}
data2 <- rbind( 
  cbind( rep(0, 100), rep(0,  100), rep(0, 100), rep(0, 100)),
  cbind( rep(0, 200), rep(1,  200), rep(0, 200), rep(0, 200)),
  cbind( rep(1,  40), rep(0,   40), rep(0,  40), rep(0,  40)),
  cbind( rep(1,1280), rep(1, 1280), rep(0,1280), rep(0,1280)),
  cbind( rep(0, 200), rep(0,  200), rep(1, 200), rep(0, 200)),
  cbind( rep(0, 100), rep(1,  100), rep(1, 100), rep(0, 100)),
  cbind( rep(1,  50), rep(0,   50), rep(1,  50), rep(0,  50)),
  cbind( rep(1, 200), rep(1,  200), rep(1, 200), rep(0,  200)),
  cbind( rep(0,  80), rep(0,   80), rep(0,  80), rep(1,  80)),
  cbind( rep(0, 100), rep(1,  100), rep(0, 100), rep(1, 100)),
  cbind( rep(1,  20), rep(0,   20), rep(0,  20), rep(1,  20)),
  cbind( rep(1, 250), rep(1,  250), rep(0, 250), rep(1, 250)),
  cbind( rep(0, 260), rep(0,  260), rep(1, 260), rep(1, 260)),
  cbind( rep(0, 100), rep(1,  100), rep(1, 100), rep(1, 100)),
  cbind( rep(1,  50), rep(0,   50), rep(1,  50), rep(1,  50)),
  cbind( rep(1, 625), rep(1,  625), rep(1, 625), rep(1, 625)))
data2 <- as.data.frame(data2)
names(data2) <- c("End.Ca", "Estr", "HxGall", "Hyper") 
data2[,names(data2)] <- lapply(data2[,names(data2)] , factor)
table(data2)
```
#### Full model
```{r}
mod.full2 <- glm( End.Ca ~ Estr*HxGall*Hyper,
                 data=data2, family=binomial)
summary(mod.full2)
plot(allEffects(mod.full2), multiline=TRUE)
```

#### ORs: reference group is women without ERT and no hypertension and no history of gall bladder disease
```{r}
OR100 <- exp(coef(mod.full2)[2])
OR110 <- exp(coef(mod.full2)[2]+coef(mod.full2)[3]+coef(mod.full2)[5])
OR101 <- exp(coef(mod.full2)[2]+coef(mod.full2)[4]+coef(mod.full2)[6])
OR111 <- exp(coef(mod.full2)[2]+coef(mod.full2)[3]+coef(mod.full2)[4]+
         coef(mod.full2)[5]+coef(mod.full2)[6]+coef(mod.full2)[7]+
         coef(mod.full2)[8])
OR100
OR110
OR101
OR111
```


