---
title: "BS852 - Class 7 FHS"
author: ""
date: ""
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

knitr::opts_knit$set(root.dir = "C:\\Users\\BS852\\Class 7")

library(survival)
```

## Analysis using the Framingham data set

### Prepare data
```{r}
framdat3 <- read.csv("framdat3.csv", header=T, na.strings=c("."))
summary(framdat3)
framdat3 <- na.omit(framdat3)
framdat3 <- subset(framdat3, framdat3$SEX == 2)

## 1499 women with complete data
dim(framdat3)

## 291 women who developed CHD 
length(which(framdat3$CHDyear > 0))

## 51 women with glucose intolerance
length(which(framdat3$GLI > 0))

## 22 women with glucose intolerance who developed CHD 
length(which(framdat3$CHDyear > 0 & framdat3$GLI > 0))

## 269 women without glucose intolerance who developed CHD 
length(which(framdat3$CHDyear > 0 & framdat3$GLI == 0))

## this is the event -> 1 if developed CHD, else 0
framdat3$chd_sw <- as.numeric(framdat3$CHDyear > 0)  

```  

### Code chdtime (follow-up time) variable

#### The variable of interest is time of CHD. We code chdtime as time of occurrence of disease or death.
#### We consider survival in 24 years from baseline, when the test for glucose intolerance was done.
#### We start by setting chdtime = 24 (maximum time in years from baseline) then, for people with the event (chd_sw == 1) the time of CHD event is CHDyear
```{r}
framdat3$chdtime = rep(24, length(framdat3$chd_sw))

framdat3$chdtime[ which(framdat3$chd_sw == 1)] <- framdat3$CHDyear[ which(framdat3$chd_sw == 1)]
```

#### Censored observations are all cases who are disease free at the end of the study or who died after baseline 
```{r} 
framdat3$chdtime[ which(framdat3$DTHyear > 0 & framdat3$chd_sw == 0)] <-
  framdat3$DTHyear[ which(framdat3$DTHyear > 0 & framdat3$chd_sw == 0)]
```
 
### KM Curves and Log-Rank Test
```{r}
fit.2 <- survfit(Surv(chdtime, chd_sw) ~ GLI, data=framdat3)
summary(fit.2)

# Kaplan-Meier Plot
plot(fit.2, mark.time=T, mark=c(1,2), col=c(1,2), lwd=2, ylim=c(0,1),
             xlab="Time (years)", ylab="Disease free survival", cex.axis=1.5, cex.lab=1.5)
legend(x=1, y=0.40, legend=c("No GLI","GLI"),
             col=c(1,2), lwd=2, cex=1.2)

# Log-Rank Test
survdiff(Surv(chdtime, chd_sw) ~ GLI, data=framdat3)
```

### A fancier survival plot using the **survminer** package
#### Reference https://rpkgs.datanovia.com/survminer/index.html
```{r warning=FALSE}
# A fancier survival plot using the 
library(survminer)
ggsurvplot(
  fit.2, 
  data = framdat3,
  xlab="Time (years)",
  size = 1,                 # change line size
  palette =
    c("#FF3333","#0066CC"),       # custom color palettes
  conf.int = TRUE,          # Add confidence interval
  pval = TRUE,              # Add p-value
  risk.table = TRUE,        # Add risk table
  risk.table.col = "strata",# Risk table color by groups
  legend.labs =
    c("GLI=0", "GLI=1"),    # Change legend labels
  risk.table.height = 0.3,  # Useful to change when you have multiple groups
  ggtheme = theme_bw()      # Change ggplot2 theme
)

```
 
