---
title: 'Class 5: Logistic Regression'
author: ""
date: ""
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# set working directory if necessary:
knitr::opts_knit$set(root.dir = "C:\\")
```

#### This chunk of code 
* reads data, 
* extract records with complete data
* selects records for *FEMALES* and those with *CHD=0 OR CHD>4*
```{r }
   
framdat2 <- read.table("framdat2.txt",header=T, na.strings=c("."))
names(framdat2)

work.data <- na.omit(framdat2[,-c(3,4)]) ## drop DTH and CAU columns and remove rows with missing data 
work.data <- subset(work.data, (work.data$SEX == 2) & (work.data$CHD==0 | work.data$CHD > 4))
work.data$chd_sw = work.data$CHD >= 4  ## 1 is event, 0 is no event
``` 

#### CRUDE ANALYSIS
We fit the model with only GLI, using the glm() function. "family=binomial" indicates that this is a logistic regression model.
```{r}

mod.crude <- glm(chd_sw ~ GLI, family=binomial, data=work.data)
summary(mod.crude)

```

##### Output for ORs, and Wald tests using the *aod* package
```{r }     
confint.default(mod.crude)
exp(cbind(OR = coef(mod.crude), confint.default(mod.crude))) 
library(aod) # inlcudes function wald.test()
wald.test(b = coef(mod.crude), Sigma = vcov(mod.crude), Terms = 2)
```

#### ADJUSTED ANALYSIS
model with only confounder
```{r }   
mod.age <- glm(chd_sw ~ AGE, family=binomial, data=work.data)
summary(mod.age)
```

#### model fit and summary of parameter estimates - adjusted model 
```{r }
mod.age.adjusted <- glm(chd_sw ~ GLI + AGE, family=binomial, data=work.data)
summary(mod.age.adjusted)
summary(mod.age.adjusted)$coefficients
```

##### to produce the LRT, use the anova function
```{r }
anova(mod.age, mod.age.adjusted)
```

##### estimate of adjusted OR and confidence intervals
```{r }
# regression coefficients
confint.default(mod.age.adjusted)

print( "Adjusted OR and 95% CI")
exp(cbind(OR = coef(mod.age.adjusted), confint.default(mod.age.adjusted)))  
wald.test(b = coef(mod.age.adjusted), Sigma = vcov(mod.age.adjusted), Terms = 2)
```

##### Question: is there confounding?
```{r }
exp(coef(mod.crude)["GLI"])/exp(coef(mod.age.adjusted)["GLI"])
```

#### Interpretation of fitted values
```{r }
log.odds.CHD <- log(mod.age.adjusted$fitted.values/(1-mod.age.adjusted$fitted.values))
plot(work.data$AGE,log.odds.CHD,xlab="Age", ylab="log-odds-CHD")
points(work.data$AGE[work.data$GLI==1],log.odds.CHD[work.data$GLI==1],col=2)

odds.CHD <- mod.age.adjusted$fitted.values/(1-mod.age.adjusted$fitted.values)
plot(work.data$AGE,odds.CHD,xlab="Age", ylab="odds-CHD")
points(work.data$AGE[work.data$GLI==1],odds.CHD[work.data$GLI==1],col=2)
   
```

####  model with multiple confounders -- model fit and summary of parameter estimates
```{r }
mod.adjusted <- glm(chd_sw ~ GLI + AGE + CSM + FVC + MRW + SPF, family=binomial, data=work.data)
summary(mod.adjusted)
exp(cbind(OR = coef(mod.adjusted), confint.default(mod.adjusted)))  
wald.test(b = coef(mod.adjusted), Sigma = vcov(mod.adjusted), Terms = 3)
```

##### fit model with only confounders   
```{r }
mod.confounders <- glm(chd_sw ~ AGE + CSM + FVC + MRW + SPF, family=binomial, data=work.data)
summary(mod.confounders)
anova(mod.confounders,mod.adjusted)
```
