---
title: "BS852 Class 4"
author: ""
date: ""
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

knitr::opts_knit$set(root.dir =  "C:\\Users\\BS852_Fall2020\\Class 4\\R & SAS")
```

## Matched Pairs

### Import data
```{r}
pairs <- read.table("matched-pairs.txt", col.names=c("match","exposed","diseased"))
```

#### 1 row per *individual*
```{r}
head(pairs) 
```

#### Create labels
```{r}
pairs$exposed <- factor(pairs$exposed, levels = c(1,0), labels = c("Exposed", "Not Exposed"))
pairs$diseased <- factor(pairs$diseased, levels = c(1,0), labels = c("Disease", "No disease"))
table(pairs$exposed,pairs$diseased)
```

### Matched pairs can be analyzed using the Mantel-Haenszel method. Each matched pair is treated as a separate stratum 
```{r}
mantelhaen.test(pairs$exposed,pairs$diseased,pairs$match, correct=FALSE)
```

```{r}
mantelhaen.test(pairs$exposed,pairs$diseased,pairs$match)
```

### Set up data to use McNemar's test: a table of *pairs*
#### exposed.d is 0 (No Disease) or 1 (Disease), and unexposed.d is 0 (No Disease) or 1 (Disease) 
```{r }
exposed.d   <- c( rep("Disease", 15), rep("No Disease",35))
unexposed.d <- c( rep("Disease",5),   rep("No Disease",10), 
                  rep("Disease",5),   rep("No Disease",30))

# table of pairs:
table(exposed.d , unexposed.d)
```

### Analyze data with mcnemar.test()
```{r}
mcnemar.test(table(exposed.d , unexposed.d), correct=FALSE) 
```

```{r}
mcnemar.test(table(exposed.d , unexposed.d))
```

## R to 1 matching

### Import data
```{r } 
Rto1 <- read.csv("Rto1matching.csv")
```

#### 1 row per *individual*
```{r}
head(Rto1) 
```

#### Create labels
```{r}
Rto1$casecon <- factor(Rto1$casecon, levels = c(1,0), labels = c("Case (bladder cancer)", "Control"))
Rto1$smoke <- factor(Rto1$smoke, levels = c(1,0), labels = c("> pack a day", "< pack a day"))
table(Rto1$smoke,Rto1$casecon)
```

### Matched sets can be analyzed using the Mantel-Haenszel method. Each matched set is treated as a separate stratum 
```{r}
mantelhaen.test(Rto1$smoke, Rto1$casecon, Rto1$casenum, correct=FALSE)
```

```{r}
mantelhaen.test(Rto1$smoke, Rto1$casecon, Rto1$casenum)
```
