---
title: "HW10_852_EpiStats"
output: pdf_document
date: "2022-12-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(epitools)
library(aod) # For Wald tests
library(MASS) # for glm.nb()
library(lmtest) # for lrtest()
library(gee) # for gee()
library(geepack) # for geeglm()
```

```{r data}
skin <- read.csv("/home/elkip/Datasets/skinca.csv")
head(skin)
```

## Question 1

### Introduction

The investigation is observing the relationship between rural vs urban setting and skin cancer using a Poisson regression. Data was collected from the Massachusetts cancer registry and the Massachusetts Census to get data on the incidence of skin cancer over a 4 year peroid by town and population information.

The variables in this data set are

1.  town/city name;

2.  sex, coded 1 for females, 0 for males;

3.  an indicator coded 1 for rural towns, 0 for urban towns;

4.  the sex-specific town population (population of females for lines relating to females, population of males for lines relating to males), in 1,000 of people;

5.  the sex-specific number of cancers from the town.

### 1a. Preliminary analysis

-   Find the total number of cancers and the total population size for the rural towns and urban towns, ignoring gender.

-   Calculate the 4-year cancer rate per 1000 population for urban towns and for rural towns

-   Give a confidence interval for each of these rates

-   Calculate the incidence rate ratio comparing the cancer rates for rural vs urban towns (no CI)

    ------------------------------------------------------------------------

```{r preliminary}
# Total number of cancers and population
rural = skin[which(skin$rural == 1),]
urban = skin[which(skin$rural == 0),]
tot_cancer.rural <- sum(rural$cases)
tot_pop.rural <- sum(rural$population)*1000
tot_cancer.urban <- sum(urban$cases)
tot_pop.urban <- sum(urban$population)*1000
print(paste("Total cancers rural: ", tot_cancer.rural))
print(paste("Total population rural: ", tot_pop.rural))
print(paste("Total cancers urban: ", tot_cancer.urban))
print(paste("Total population urban: ", tot_pop.urban))

# Four Year cancer rate per 1000 population (d)
rate.rural <- (tot_cancer.rural / tot_pop.rural) * 1000
rate.urban <- (tot_cancer.urban / tot_pop.urban) * 1000
print(paste("4 Year Cancer Rate Rural: ", rate.rural, " per 1000 persons"))
print(paste("4 Year Cancer Rate Urban: ", rate.urban, " per 1000 persons"))

# CI For Rate 
# (1000 / n) (d +/- (1.96 x square root of d)
se.rural <- sqrt(tot_cancer.rural)
se.urban <- sqrt(tot_cancer.urban)
rate_ci_u.rural <- (tot_cancer.rural + 1.96*se.rural)*(1000/tot_pop.rural)
rate_ci_l.rural <- (tot_cancer.rural - 1.96*se.rural)*(1000/tot_pop.rural)
rate_ci_u.urban <- (tot_cancer.urban + 1.96*se.urban)*(1000/tot_pop.urban)
rate_ci_l.urban <- (tot_cancer.urban - 1.96*se.urban)*(1000/tot_pop.urban)
print(paste("4 Year Cancer Rate 95% CI Rural: ", rate_ci_l.rural, ", ", rate_ci_u.rural))
print(paste("4 Year Cancer Rate 95% CI Urban: ", rate_ci_l.urban, ", ", rate_ci_u.urban))

# Incidence rates rural vs urban
rate.ratio <- rate.rural / rate.urban
print(paste("Incidence rate ratio: ", rate.ratio))
```

### 1b. Poisson regression crude

Fit a Poisson regression model predicting the rate of skin cancer cases from the rural / urban variable, using the log population as the 'offset' (so you will be analyzing cases per 1000 population)

-   What is the rate ratio for skin cancer for rural vs urban people, and how does the rate ratio from this Poisson model compare to the incidence rate ratio you calculated in 1a?
-   From the Poisson model give a 95% CI for this rate ratio.
-   Are the rates of skin cancer significantly different between rural towns and urban towns?

------------------------------------------------------------------------

```{r}
# Poisson reg
poi.crude <- glm(cases ~ rural, family = poisson(link="log"), offset = log(population), data = skin)
summary(poi.crude)

# Rate ratio and 95% CI
exp(cbind(IRR = coef(poi.crude), confint.default(poi.crude))) 
```

The rate ratio for rural vs urban is 1.98 (95% CI: 1.43, 2.73), this is the same as the IRR I calculated above. It would appear that the rate of cancer is significantly higher in those who live in rural areas than urban areas.

### 1c. Poisson regression adjusted for sex

Fit a Poisson model predicting the risk of skin cancer for rural vs. urban people, adjusting for sex.

-   What is the rate ratio for skin cancer for rural vs. urban people after adjusting for sex?
-   From the Poisson model, give a 95% CI for this rate ratio. Is sex a confounder here? Provide evidence to support your conclusion.

------------------------------------------------------------------------

```{r}
# Adjusted Poisson model
poi.adj <- glm(cases ~ rural + sexf, family = poisson(link="log"), offset = log(population), data = skin)
summary(poi.adj)

# IRR and 95% CI 
exp(cbind(IRR = coef(poi.adj), confint.default(poi.adj)))

# Test for confounding in beta (crude - adjusted) / crude
print(paste("Percent diff in rate: ", abs(.6810 - .6802) / .6810))
```

The rate ratio for skin cancer in rural vs urban settings adjusted for sex is .6802 (95% CI: 1.43, 2.70). The difference in the crude and adjusted rate is .11%, which is less than 10% so we fail to identify sex as a confounder in the relationship between rural vs urban setting and cancer risk.

### Conclusion

H~0~: The expected rate of cancer in rural and urban areas is the same.

H~a~: Not H0

In the preliminary analysis we observed cancer rate in rural areas was 1.55 per 1000 persons (95% CI: .1.1, 2.04) which is much higher compared to urban areas where is was only .79 per 1000 persons (95% CI: .72, .85). This results in an IRR of 1.98, implying the expected number of cancer in rural areas is greater than that of urban areas by a factor of 1.98. The Poisson regression estimated the effect is .6810 (z = 4.127, p \< 0001), which reflects the same risk ratio of 1.98. Thus there is evidence to reject the null hypothesis, and conclude there is a significant association between urban and rural settings and the expected rate of cancer. Also we did not find evidence of confounding when adjusted for sex.

## Question 2

```{r}
hypo <- read.csv("/home/elkip/Datasets/hypoglycemia.episodes.csv")
head(hypo)
```

### Introduction

127 patients at a clinical practice with type I diabetes were monitored with a continuous glucose monitor (CGM) all day for 7 days. The aim is to study the number of hypoglycemic episodes which occur (blood glucose \< 70mg/dL).

Other factors recorded:

-   total length of time CGM was function

-   gender

-   age

-   years since diagnosis

-   if the patient already owned a CGM

### 2a. Poisson Regression

```{r}
hypo.poi <- glm(hypoepisodes ~ female + age + yearssincedx + ownmonitor, family = poisson(link="log"), offset = log(days), data = hypo)
summary(hypo.poi)
```

log(lambda) = -.132 + .009\*female + .106\*age - .01\*years_since_dx - .209\*monitor

The outcome is the log of the event rate of hypoglycemic episodes. The intercept represents a decrease in log(lambda) of -.132 for the baseline rate (when all predictors are 0).

For females the estimated number of hypoglycemic episodes increases by a factor of e\^.00937 = 1.009516 compared to males.

For each additional year of age the estimated number of hypoglycemic episodes increases by a factor of e\^.015628 = 1.015751.

For each additional year since diagnosis the estimated number of hypoglycemic episodes decreases by a factor of e\^-0.010213 = 0.989839.

For patients that already owned a monitor the estimated number of hypoglycemic episodes decreases by a factor of e\^-0.208521 = 0.811784.

### 2b. Patients with high number of hypoglycemic episodes?

If the clinic wants to target patients with relatively high numbers of hypoglycemic episodes, in order to educate them on best practices to reduce their numbers of hypoglycemic episodes, whom should they target? Why?

------------------------------------------------------------------------

We observe in the regression above that owning a monitor is significantly associated with lower rates of hypoglycemic episodes compared to those who don't own a monitor when adjusting for age, sex and years since diagnosis. Thus, the researcher should target patients with type I diabetes who did not own a monitor prior to the study.

### 2c. Which variables are significantly associated with lower numbers of hypoglycemic episodes?

Find the 95% confidence interval for the effect size for each such variable.

------------------------------------------------------------------------

In the above Poisson regression the 'monitor' variable, if the patient already owned a monitor, was the only significantly associated with lower numbers of hypoglycemic episodes.

```{r}
cbind(beta = coef(hypo.poi), confint.default(hypo.poi))
```

The 95% Confidence interval for the coefficient of the monitor variable is (-.37509, -.04194). Since it does not contain zero it gives evidence at alpha .05 to reject the null hypothesis.

### 2d. Do gender, age, and years since diagnosis as a group confound the association between hypoglycemic episodes and owning CGM monitors?

```{r}
hypo.poi_crude <- glm(hypoepisodes ~ ownmonitor, family = poisson(link="log"), offset = log(days), data = hypo)
summary(hypo.poi_crude)
print(paste("Prop diff in coeff: ", abs((-.23728 - -.20852) / -.23728)))
```

We observe a 12.1% difference in the coefficient of the monitor variable between crude and sex, years_since_dx, and age adjusted model. Since this is greater than 10% we conclude the association between owning a monitor and the event rate of hypoglycemic episodes is confounded by sex, years_since_dx, and age.

### Conclusion

H~0~: The expected rate of hypoglycemic episodes in patients with Type I diabetes is the same across those who did and didn't use a glucose monitor, adjusted for sex, age, and the number of years since diagnosis as a group.

H~a~: Not H0

The Poisson estimated the coefficient of the monitor variable to be -.20851, meaning those patients who used a monitor have an expected number of hypoglycemic episodes which is lower by factor of .8118 compared to those who didn't wear a monitor. This variable was the only significant coefficient in the model with (z=-2.454, p\< .0141). It was observed when gender, age and years since diagnosis were removed from the model the change in the coefficient of the monitor variable change by more than 10%, suggesting that the association between owning a monitor and the event rate of hypoglycemic episodes is confounded by sex, years_since_dx, and age as a group. Thus, we conclude the best model includes all the variables given and there is evidence to reject the null hypothesis and conclude that the expected rate of hypoglycemic episodes in patients with Type I diabetes is NOT the same across those who did and didn't use a glucose monitor, adjusted for sex, age, and the number of years since diagnosis as a group.
