---
title: "Homework 1"
author: "Mitchell Henschel"
date:   "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: united
    code_folding: show
    toc: true
    style: BS831.css
    toc_float: true
---

## Settings

```{r settings, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)

library(Biobase)
library(BS831)

OMPATH <- Sys.getenv("OMPATH")
if ( OMPATH=="" ) stop( "OMPATH must be set" )
print(OMPATH) 
DATA <- Sys.getenv("DATAPATH")
```

# Exercise 1: Creating an ExpressionSet from multiple parts

```{r}
EXP <- as.matrix(read.table(file.path(DATA, "hw1/ESet_exprs.xls"), sep = '\t', check.names = FALSE))
PDAT <- read.table(file.path(DATA, "hw1/ESet_pData.xls"), sep = '\t', header = TRUE)
FDAT <- read.table(file.path(DATA, "hw1/ESet_fData.xls"), sep = '\t', header = TRUE)

ESET <- ExpressionSet(assayData = EXP, phenoData = AnnotatedDataFrame(PDAT),
                               featureData = AnnotatedDataFrame(FDAT))

gMedian <- rowMedians(exprs(ESET))
sMedian <- rowMedians(t(exprs(ESET)))
```

```{r}
print(ESET)
```

```{r}
exprs(ESET)[1:5,1:5]
```

```{r}
head(gMedian)
```

```{r}
head(sMedian)
```

# Exercise 2: Median vs. MAD plot

```{r}
MED <- apply(exprs(ESET), 1, median)
MAD <- apply(exprs(ESET), 1, mad)

# Combine to matrix and take top 10 rows by order of MAD
mat <- cbind(MED, MAD)
mat1 <- mat[order(mat[,2], decreasing = TRUE),][1:100,]
plot(MED,MAD,pch=20)
points(mat1, pch=1, col='red')
legend(x = "topright", legend = c("all genes", "top 100 genes by MAD"), col = c('black', 'red'), pch = c(20,1))
```

```{r}
print(mat1[1:10,])
```

Exercise 3: "Genotype" - Phenotype Association

```{r}
top <- ESET$pam50
boxplot(exprs(ESET)[1,] ~ ESET$pam50,
        xlab = "pam50 type", ylab = "SCGB2A2",
        col = c('red','green','blue','purple'))

```

SCGB2A2 appears to be a good marker for LumA, as it has the highest association and low variance.

# Exercise 4: Bayes Thereom

Suppose a new HIV test is claimed to have "95% sensitivity and 98% specificity." In a population with an HIV prevalence of 1/1000, what is the probability that a patient testing positive actually has HIV?

```{r}
ANS <- (.95 * 1/1000)/(.95* 1/1000 + (1 - .98)*(1- 1/1000))
```

```{r}
print(paste(ANS*100, "%", sep=""))
```

There is a 4.5% chance the patient who tests positive for HIV actually has HIV.

# Exercise 5: Matrix Operations

```{r}
set.seed(143)
X <- cbind(rep(1, 50), rnorm(50, 0, 1), rnorm(50, 0, 1))
y <- 3 + -4*X[,2] + 2*X[,3] + rnorm(50, 0, 1) # y = 3 + -4(x_1) + 2(x_2) + E
```

```{r}
BetaEstimator <- function(X, y){
  Beta <- solve(t(X)%*%X)%*%t(X)%*%y
  return(Beta)
}
```

```{r}
BetaEstimator(X, y)
```
