---
title: "Homework 2"
author: "Mitchell Henschel"
date:   "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: united
    code_folding: show
    toc: true
    style: BS831.css
    toc_float: true
---

## Settings

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)

library(Biobase)
library(limma)
library(BS831)

OMPATH <- Sys.getenv("OMPATH")
if ( OMPATH=="" ) stop( "OMPATH must be set" )
print(OMPATH) 
DATA <- Sys.getenv("DATAPATH")
```

## Upload and Inspect Data

```{r}
data(dm10)
tail(pData(dm10))
```

Show chemicals:

```{r}
head(table(pData(dm10)$CHEMICAL))
```

Show count samples by carcinogenicity category:

```{r}
table(pData(dm10)$Carcinogen_liv, useNA="ifany")
```

Simplify the phenotype name for easier handling:

```{r}
colnames(pData(dm10)) <- gsub("Carcinogen_liv", "carc", colnames(pData(dm10)))
hist(exprs(dm10))
```

Restrict the dataset to samples needed for the analysis (exclude controls and un-annotated samples)

```{r}
# Extact treated samples
dat <- dm10[,pData(dm10)$carc %in% c('NON-CARC', 'CARCINOGEN')]
# Exclude samples wit GenTox' NA's
dat <- dat[,!is.na(pData(dat)$GenTox)]
# Order by phenotype labels
data <- dat[,order(dat$carc)]
table(dat$carc)
```

## Variation Filtering

Reduce the number of genes to analyze by variation filtering. Use a MAD filter to reduce the number of genes to 2000.

```{r}
MED <- apply(exprs(dat), 1, median)
MAD <- apply(exprs(dat), 1, mad)
top.genes <- order(MAD, decreasing = TRUE)[1:2000]

plot(MED, MAD, pch=".", log="")
points(MED[top.genes], MAD[top.genes], pch=".", col="red")
legend("topright", pch=20, col = c('black', 'red'), legend = c("all genes", "genes passing MAD filter"))
```

```{r}
# Assign data to the filtered data
dat <- dat[top.genes,]
```

# Exercise 1
### Diffanal with t.test - unequal variance
```{r}
ttestRes0 <- t.test(rownames(pData(dat))~pData(dat)$carc, var.equal = FALSE)
```

# Exercise 2
### Diffanal with t.test - equal variance

```{r}

```

# Exercise 3
### Diffanal with lm - no confounders

```{r}

```

# Exercise 4
### Controlling for Confounders (Extra Credit)

```{r}

```

# Exercise 5

### Diffanal using limma (no confounders)

```{r}

```

# Exercise 6

### Controlling for Confounders (Extra Credit)

```{r}

```

# Exercise 7

### Data Acquisition to Differently Expressed Genes
