---
title: "Homework 5"
author: "Mitchell Henschel"
date:   "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: united
    code_folding: show
    toc: true
    style: BS831.css 
    toc_float: true
---

## Settings
```{r settings, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
require(GEOquery)
require(Biobase)
require(dplyr)
require(GSVA)
require(pheatmap)
library(BS831)
OMPATH <- Sys.getenv("DATAPATH")
```

## Loading the Data
```{r}
LOAD <- readRDS(file.path(OMPATH,"LOAD1.RDS"))
print(dim(LOAD))
```
```{r}
## show the phenotype annotation and brain regions
table(LOAD$characteristics_ch2,LOAD$description)
```
## Cleaning the Data
We first remove probes w/o associated gene symbols. We then simplify the phenotype annotation by naming it loadStatus. Finally, we remove gene replicates (by picking the replicate w/ highest variation), and remove the genes with NA’s.
```{r}
## restrict to probes with mapping gene symbols
LOAD1 <- LOAD[!is.na(fData(LOAD)$gene_symbol) & fData(LOAD)$gene_symbol!="",]; nrow(LOAD1)
```
```{r}
## let us make the phenotype annotation more intuitive
colnames(pData(LOAD1)) <- gsub("characteristics_ch2","loadStatus",colnames(pData(LOAD1)))
LOAD1$loadStatus <- factor(c("control","load")[as.numeric(LOAD1$loadStatus=="disease: A")+1],
                           levels=c("load","control"))
print(table(LOAD1$loadStatus,LOAD1$description))
```
```{r}
## some gene symbols have multiple entries
c(total=nrow(LOAD1),unique=length(unique(fData(LOAD1)$gene_symbol)))
```
```{r}
## we 'uniquefy' by taking the replicate w/ largest variation
## (one can also use dplyr to do the same perhaps more elegantly)
SD <- apply(exprs(LOAD1),1,sd,na.rm=TRUE)              # compute SD of each gene
LOAD1 <- LOAD1[order(SD,decreasing=TRUE),]             # sort genes by SD (largest to smallest)
LOAD1 <- LOAD1[match(unique(fData(LOAD1)$gene_symbol), # pick first occurrence of each
                     fData(LOAD1)$gene_symbol),]
featureNames(LOAD1) <- fData(LOAD1)$gene_symbol
c(total=nrow(LOAD1),unique=length(unique(fData(LOAD1)$gene_symbol)))
```
```{r}
## remove probes w/ NA's (alternative would be to impute them)
LOAD2 <- LOAD1[!apply(is.na(exprs(LOAD1)),1,any),]
c(missing.LOAD1=sum(is.na(exprs(LOAD1))),missing.LOAD2=sum(is.na(exprs(LOAD2))))
```
## APEO Signature Projection
```{r}
apoeSIG <- list(APOEup=toupper(c("BIRC2","CEP57","VPS29","PSME1","TBCA","UBA2","KMT2C","KIN","CKAP2")),
                APOEdn=toupper(c("S100A13","LRRN1","APOE","C5orf38","CTF1","APOB","CRYZL1")))

## let's check overlap of the signature with genes in the LOAD dataset
data.frame(
    original=sapply(apoeSIG,length),
    matching=sapply(lapply(apoeSIG,intersect,toupper(featureNames(LOAD2))),length)
) %>% dplyr::mutate(delta=original-matching)
```
# Exercise 1: GSVA Projection
Here, you are asked to perform GSVA-based signature projection on the LOAD dataset based on the defined apoe signatures (APOEup and APOEdn). You are also askded to compute the final ‘UP-DN’ APOE score as described in Figure 2 of Sebastiani et al.
```{r}
## running gsva (with mx.diff=FALSE)
LOADgsva <- ## perform gsva

## adding combined up-dn scores
LOADgsva <- rbind(LOADgsva,
                  APOEscore= ...) ## compute combined 'up - down' score

## we show few entries of the resulting expression matrix
print(LOADgsva[,1:5])
```

# Exercise 2: APOE Signature Projection Heatmap
Here you are asked to generate a heatmap of the enrichment scores, with the columns ordered by the ‘UP-DN’ score, and with the addition of the phenotypic annotation of each column by LOAD status.
```{r}
## just checking
all(sampleNames(LOAD2)==colnames(LOADgsva))

## plot heatmap (w/ pheatmap or ComplexHeatmap)
```

# Exercise 3: Test Association by KS
Here, you are asked to test by Kolmogorov-Smirnov Test whether the LOAD patients are skewed towards higher APOE scores.
```{r}
all(colnames(LOADgsva)==sampleNames(LOAD2)) ## just checking

apoe <- LOADgsva["APOEscore",]
loadRnk <- ## rank by APOEscore

## compute KS score (use function R/ksGenescore.R)
ksGenescore(...)
```

# Exercise 4: Test Association by Differential Analysis
We then test by wilcox and anova tests whether there is a significant difference between the distribution of scores across regions (wilcox) and within regions (anova)
```{r}
## across regions
pval <- wilcox.test(...)
boxplot(...) # or ggplot if you prefer

## within regions
pval <- anova(...)
boxplot(...) # or ggplot if you prefer
```
