---
title: "Homework 5"
author: "Mitchell Henschel"
date:   "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: united
    code_folding: show
    toc: true
    style: BS831.css 
    toc_float: true
---

## Settings
```{r settings, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
require(GEOquery)
require(Biobase)
require(dplyr)
require(GSVA)
require(pheatmap)
library(BS831)
OMPATH <- Sys.getenv("DATAPATH")
```

## Loading the Data
```{r}
LOAD <- readRDS(file.path(OMPATH,"LOAD1.RDS"))
print(dim(LOAD))
```
```{r}
## show the phenotype annotation and brain regions
table(LOAD$characteristics_ch2,LOAD$description)
```
## Cleaning the Data
We first remove probes w/o associated gene symbols. We then simplify the phenotype annotation by naming it loadStatus. Finally, we remove gene replicates (by picking the replicate w/ highest variation), and remove the genes with NA’s.
```{r}
## restrict to probes with mapping gene symbols
LOAD1 <- LOAD[!is.na(fData(LOAD)$gene_symbol) & fData(LOAD)$gene_symbol!="",]; nrow(LOAD1)
```
```{r}
## let us make the phenotype annotation more intuitive
colnames(pData(LOAD1)) <- gsub("characteristics_ch2","loadStatus",colnames(pData(LOAD1)))
LOAD1$loadStatus <- factor(c("control","load")[as.numeric(LOAD1$loadStatus=="disease: A")+1],
                           levels=c("load","control"))
print(table(LOAD1$loadStatus,LOAD1$description))
```
```{r}
## some gene symbols have multiple entries
c(total=nrow(LOAD1),unique=length(unique(fData(LOAD1)$gene_symbol)))
```
```{r}
## we 'uniquefy' by taking the replicate w/ largest variation
## (one can also use dplyr to do the same perhaps more elegantly)
SD <- apply(exprs(LOAD1),1,sd,na.rm=TRUE)              # compute SD of each gene
LOAD1 <- LOAD1[order(SD,decreasing=TRUE),]             # sort genes by SD (largest to smallest)
LOAD1 <- LOAD1[match(unique(fData(LOAD1)$gene_symbol), # pick first occurrence of each
                     fData(LOAD1)$gene_symbol),]
featureNames(LOAD1) <- fData(LOAD1)$gene_symbol
c(total=nrow(LOAD1),unique=length(unique(fData(LOAD1)$gene_symbol)))
```
```{r}
## remove probes w/ NA's (alternative would be to impute them)
LOAD2 <- LOAD1[!apply(is.na(exprs(LOAD1)),1,any),]
c(missing.LOAD1=sum(is.na(exprs(LOAD1))),missing.LOAD2=sum(is.na(exprs(LOAD2))))
```
## APEO Signature Projection
```{r}
apoeSIG <- list(APOEup=toupper(c("BIRC2","CEP57","VPS29","PSME1","TBCA","UBA2","KMT2C","KIN","CKAP2")),
                APOEdn=toupper(c("S100A13","LRRN1","APOE","C5orf38","CTF1","APOB","CRYZL1")))

## let's check overlap of the signature with genes in the LOAD dataset
data.frame(
    original=sapply(apoeSIG,length),
    matching=sapply(lapply(apoeSIG,intersect,toupper(featureNames(LOAD2))),length)
) %>% dplyr::mutate(delta=original-matching)
```
# Exercise 1: GSVA Projection
Here, you are asked to perform GSVA-based signature projection on the LOAD dataset based on the defined apoe signatures (APOEup and APOEdn). You are also askded to compute the final ‘UP-DN’ APOE score as described in Figure 2 of Sebastiani et al.
```{r}
## running gsva (with mx.diff=FALSE)
LOADgsva_full <- gsva(LOAD2, apoeSIG, method = "gsva", mx.diff = FALSE)

## adding combined up-dn scores
LOADgsva <- rbind(exprs(LOADgsva_full),
                  APOEscore= exprs(LOADgsva_full)[1,] - exprs(LOADgsva_full)[2,], loadstatus = factor(c("control","load")[as.numeric(LOAD2$loadStatus)],
                           levels=c("load","control")))

## we show few entries of the resulting expression matrix
print(LOADgsva[,1:5])
```

# Exercise 2: APOE Signature Projection Heatmap
Here you are asked to generate a heatmap of the enrichment scores, with the columns ordered by the ‘UP-DN’ score, and with the addition of the phenotypic annotation of each column by LOAD status.
```{r}
## just checking
all(sampleNames(LOAD2)==colnames(LOADgsva))

LOADgsva2 <- LOADgsva[,order(LOADgsva[3,])]
## plot heatmap (w/ pheatmap or ComplexHeatmap)
pheatmap(as.matrix(LOADgsva2),
         scale = "row",
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         #col = c("Control" = "#377eb8", "AD" = "#e41a1c"),
         main = "APOE Signiature LOAD Projection",
         show_colnames = FALSE)
```

# Exercise 3: Test Association by KS
Here, you are asked to test by Kolmogorov-Smirnov Test whether the LOAD patients are skewed towards higher APOE scores.
```{r}
all(colnames(LOADgsva)==sampleNames(LOAD2)) ## just checking

apoe <- as.data.frame(LOADgsva["APOEscore",])
colnames(apoe) <- c("Score")
apoe$rank <- round(rank(apoe), 0)
apoe$ID <- rownames(apoe)
apoe <- apoe %>% filter(apoe$ID %in% colnames(LOADgsva[,LOADgsva["loadstatus",]==2,drop=FALSE]))

## compute KS score (use function R/ksGenescore.R)
ksGenescore(690, apoe$rank, do.plot = TRUE)
```

# Exercise 4: Test Association by Differential Analysis
We then test by wilcox and anova tests whether there is a significant difference between the distribution of scores across regions (wilcox) and within regions (anova)
```{r}
## across regions
pval <- wilcox.test(LOADgsva["APOEscore",] ~ LOADgsva["loadstatus",])$p.value
boxplot(LOADgsva["APOEscore",] ~ LOADgsva["loadstatus",],col = c("lightblue","pink"), ylab = "APOEscore", xlab = "Wilcox Test", main = "Across Brain Regions", names = c("control", "load"))
```

```{r}
combined_classes <- as.data.frame(cbind(paste(LOADgsva_full$description,LOADgsva_full$loadStatus),colnames(LOADgsva_full)))

colnames(combined_classes) <- c("class","ID")

scoredata <- c()
scoredata$score <- LOADgsva["APOEscore",]
scoredata$ID <- colnames(LOADgsva)

anoveData <- merge(scoredata,combined_classes, by = "ID")

results <- aov(anoveData$score ~ anoveData$class)
pval <- anova(results)["Pr(>F)"][[1]][1]


anoveData$class <- factor(anoveData$class, levels = c("cerebellum load", "cerebellum control", "DLPFC load", "DLPFC control","visual cortex load", "visual cortex control"))

boxplot(anoveData$score ~ anoveData$class,col = c("pink","lightblue"), ylab = "Apoe score",xlab = NA, main = "within brain regions" )
```

