---
title: "Homework 4"
author: "Mitchell Henschel"
date:   "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: united
    code_folding: show
    toc: true
    style: BS831.css 
    toc_float: true
---

## Settings
```{r settings, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(Biobase)
library(BS831)
OMPATH <- Sys.getenv("DATAPATH")
```

## Loading the Data
```{r}
DAT <- readRDS(file.path(OMPATH,"HNSC_htseq_normalized_G1vsG3.RDS"))
dim(DAT)
```
```{r}
print(table(DAT$grade,DAT$stage))
```
```{r}
head(fData(DAT))
```
```{r}
## use hypeR helper function
HALLMARK <- hypeR::msigdb_gsets(species="Homo sapiens", category="H")
print(HALLMARK) # show few genesets (use names(HALLMARK$geneset) to see all)
```
```{r}
## extract the relevant geneset
gs <- HALLMARK$genesets[["HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION"]]
length(gs) # count the number of genes
```
```{r}
## see how many genes in the geneset are included in the dataset
length(intersect(gs,fData(DAT)$hgnc_symbol)) 
```

# Exercise 1: Testing for Enrichment
Following the examples and guidelines given in vignettes/docs/KS_enrichment, perform the enrichment test.

i. First, establish a sorting of the genes in the dataset based on the 'hi (g3) vs. low (g1) grade' phenotype.

Here, we need to carry out differential analysis comparing high grade to low grade (stored in the pData column grade). Any method of computing a differential score is acceptable, from a simple application of the t.test, to the use of packages like multtest or limma. Below we use the fast.tscore function (not shown) to compute the t-statistic for each gene.

ii. Then, determine the ranking of the genes in the geneset (gs) with respect to the sorting just calculated.

iii. Finally, use ksGenescore to test for enrichment of the geneset with respect to the phenotype 'hi (g3) vs. low (g1) grade'.

