---
title: "Homework 2"
author: "Mitchell Henschel"
date:   "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: united
    code_folding: show
    toc: true
    style: BS831.css
    toc_float: true
---

## Settings

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)

library(Biobase)
library(dplyr)
library(ggplot2)
library(tibble)
library(limma)
library(GEOquery)
library(BS831)

OMPATH <- Sys.getenv("OMPATH")
if ( OMPATH=="" ) stop( "OMPATH must be set" )
print(OMPATH) 
DATA <- Sys.getenv("DATAPATH")
```

## Upload and Inspect Data

```{r}
data(dm10)
tail(pData(dm10))
```

Show chemicals:

```{r}
head(table(pData(dm10)$CHEMICAL))
```

Show count samples by carcinogenicity category:

```{r}
table(pData(dm10)$Carcinogen_liv, useNA="ifany")
```

Simplify the phenotype name for easier handling:

```{r}
colnames(pData(dm10)) <- gsub("Carcinogen_liv", "carc", colnames(pData(dm10)))
hist(exprs(dm10))
```

Restrict the dataset to samples needed for the analysis (exclude controls and un-annotated samples)

```{r}
# Extact treated samples
dat <- dm10[,pData(dm10)$carc %in% c('NON-CARC', 'CARCINOGEN')]
# Exclude samples wit GenTox' NA's
dat <- dat[,!is.na(pData(dat)$GenTox)]
# Order by phenotype labels
data <- dat[,order(dat$carc)]
table(dat$carc)
```

## Variation Filtering

Reduce the number of genes to analyze by variation filtering. Use a MAD filter to reduce the number of genes to 2000.

```{r}
MED <- apply(exprs(dat), 1, median)
MAD <- apply(exprs(dat), 1, mad)
top.genes <- order(MAD, decreasing = TRUE)[1:2000]

plot(MED, MAD, pch=".", log="")
points(MED[top.genes], MAD[top.genes], pch=".", col="red")
legend("topright", pch=20, col = c('black', 'red'), legend = c("all genes", "genes passing MAD filter"))
```

```{r}
# Assign data to the filtered data
dat <- dat[top.genes,]
```

# Exercise 1

### Diffanal with t.test - unequal variance

```{r}
ttest_single <- t.test(exprs(dat)[1,]~dat$carc, var.equal = FALSE)
print(ttest_single)
```

```{r}
res0Wrapper <- function(gene, pheno) {
  unlist(t.test(gene ~ pheno)[c("statistic", "p.value")])
}
ttestRes0 <- data.frame(t(apply(exprs(dat), 1, res0Wrapper, pheno=dat$carc))) %>%
  mutate(q.value=p.adjust(p.value, method="BH"))
colnames(ttestRes0) <- c("t.score.t", "t.pvalue", "t.fdr")
head(ttestRes0)
```

# Exercise 2

### Diffanal with t.test - equal variance

```{r}
res1Wrapper <- function(gene, pheno) {
  unlist(t.test(gene ~ pheno, var.equal = TRUE)[c("statistic", "p.value")])
}
ttestRes1 <- data.frame(t(apply(exprs(dat), 1, res1Wrapper, pheno=dat$carc))) %>%
  mutate(q.value=p.adjust(p.value, method="BH"))
colnames(ttestRes1) <- c("t.score.t", "t.pvalue", "t.fdr")
head(ttestRes1)
```

```{r}
plot(ttestRes0[,'t.pvalue'],ttestRes1[,'t.pvalue'],pch=".",
     xlab="unequal variance",ylab="equal variance",main='t-scores comparison')
abline(0,1,col='red')
```

# Exercise 3

### Diffanal with lm - no confounders

```{r}
lm0 <- summary(lm(exprs(dat)[1,] ~ dat$carc))
print(lm0)
```

```{r}
lm.wrapper <- function(gene, pheno, confounder=NULL) {
  if (is.null(confounder)) {
    summary(lm(gene ~ pheno))$coefficients
  }
  else {
    summary(lm(gene ~ pheno + confounder))$coefficient
  }
}
lmRes0 <- data.frame(t(apply(exprs(dat), 1, lm.wrapper, pheno=dat$carc)))[, c("X6", "X8")] %>%
  rownames_to_column() %>%
  rename(lm.score=X6, lm.pvalue=X8) %>%
  mutate(lm.fdr=p.adjust(lm.pvalue, method="BH")) %>%
  column_to_rownames()
head(lmRes0)
```

```{r}
plot(ttestRes0[,"t.score.t"],lmRes0[,"lm.score"],main="compare lm to t-test w/ unequal variance")
```

```{r}
plot(ttestRes1[,"t.score.t"],lmRes0[,"lm.score"],main="compare lm to t-test w/ equal variance")
```

# Exercise 4

### Controlling for Confounders (Extra Credit)

```{r}
## count number of genotoxic samples
table(pData(dat)$GenTox,useNA="ifany")
```

```{r}
## let's see how carcinogens and genotoxicants are distributed
table(pData(dat)[,c("carc","GenTox")])
```

```{r}
lmRes1 <- data.frame(t(apply(exprs(dat), 1, lm.wrapper, pheno=dat$carc, confounder = dat$GenTox)))[, c("X8", "X11")] %>%
  rownames_to_column() %>%
  rename(lm.score=X8, lm.pvalue=X11) %>%
  mutate(lm.fdr=p.adjust(lm.pvalue, method="BH")) %>%
  column_to_rownames()
head(lmRes1)
```

```{r}
lm0.markers <- lmRes0[,"lm.fdr"]<=0.05
lm1.markers <- lmRes1[,"lm.fdr"]<=0.05
lm.markers <- lm0.markers & lm1.markers
plot(lmRes0[,'lm.score'],lmRes1[,'lm.score'],xlab="no confounder",ylab="w/ confounder")
points(lmRes0[lm0.markers,'lm.score'],lmRes1[lm0.markers,'lm.score'],col="green")
points(lmRes0[lm1.markers,'lm.score'],lmRes1[lm1.markers,'lm.score'],col="orange")
points(lmRes0[lm.markers,'lm.score'],lmRes1[lm.markers,'lm.score'],col="blue")
abline(0,1,col="red")
legend("topleft",col=c("green","orange","blue"), pch=19,
       legend=c("w/o confounders","w/ confounders","both"))
```

```{r}
maxQ <- 0.05
DFlm <- dplyr::inner_join(as.data.frame(lmRes0) %>% tibble::rownames_to_column(),
                          as.data.frame(lmRes1) %>% tibble::rownames_to_column(),
                          by="rowname",suffix=c(".0",".1")) %>%
  tibble::column_to_rownames() %>%
  dplyr::mutate(group=ifelse(lm.fdr.0<=maxQ & lm.fdr.1<=maxQ,"both",
                             ifelse(lm.fdr.0<=maxQ,"w/o confounders",
                                    ifelse(lm.fdr.1<=maxQ,"w/ confounders","neither"))))
COL <- c(neither="gray",
         "w/o confounders"="green",
         "w/ confounders"="orange",
         both="blue")
ggplot2::ggplot(DFlm,aes(x=lm.score.0,y=lm.score.1,col=group)) +
  geom_point() +
  ggplot2::geom_abline(slope=1,intercept=0,col="red") + 
  ggplot2::scale_color_manual(values=COL) +
  labs(x="w/o confounders",y="w/ confounders")
```

```{r}
apply(data.frame(unconfounded=lm0.markers,
                 confounded=lm1.markers,
                 both=lm.markers),2,sum)
```

# Exercise 5

### Diffanal using limma (no confounders)

```{r}
design <- model.matrix(~ dat$carc)
colnames(design) <- c("NON_CARC","CARC")
fit <- lmFit(dat, design)
fit <- eBayes(fit)
limmaRes0 <- topTable(fit, coef = "CARC", adjust.method="BH", n=Inf, sort.by="P")
# saveRDS(limmaRes0, file=file.path(DATA,"homework2_limmaRes0.RDS"))
head(limmaRes0)
```

```{r}
if ( any(is.na(matchIdx <- match(rownames(limmaRes0),rownames(ttestRes0)))) ) stop('mismatch')
plot(ttestRes0[matchIdx,"t.score.t"],limmaRes0[,"t"],main="compare limma to t-test w/ unequal variance")
```

```{r}
## compare to t.test (with unequal variance)
if ( any(is.na(matchIdx <- match(rownames(limmaRes0),rownames(ttestRes1)))) ) stop('mismatch')
plot(ttestRes1[matchIdx,"t.score.t"],limmaRes0[,"t"],main="compare limma to t-test w/ equal variance")
```

```{r}
## compare to lm
if ( any(is.na(matchIdx <- match(rownames(limmaRes0),rownames(lmRes0)))) ) stop('mismatch')
plot(lmRes0[matchIdx,"lm.score"],limmaRes0[,"t"],main="compare limma to lm")
```

# Exercise 6

### Controlling for Confounders (Extra Credit)

```{r}
design <- model.matrix(~ dat$carc + dat$GenTox)
colnames(design) <- c("NON_CARC","CARC", "GENTOX")
fit <- lmFit(dat, design)
fit <- eBayes(fit)
limmaRes0 <- topTable(fit, coef = "CARC", adjust.method="BH", n=Inf, sort.by="P")
# saveRDS(limmaRes0, file=file.path(DATA,"homework2_limmaRes0.RDS"))
head(limmaRes0)
```

# Exercise 7

### Data Acquisition to Differently Expressed Genes

```{r}
eset <- getGEO("GSE6264")
if (length(eset) > 1) idx <- grep("GPL1426", attr(eset, "names")) else idx <- 1
eset <- eset[[idx]]
```

```{r}
print(eset)
```

```{r}
exprs(eset)[1:5,1:5]
```

```{r}
fData(eset) <- fData(eset)[,c("ID", "Gene_Symbol")]
colnames(fData(eset)) <- c("id", "symbol")

pData(eset) <- pData(eset)[,c("geo_accession", "source_name_ch1", "characteristics_ch1")]
colnames(pData(eset)) <- c("id", "cell_line", "abuse_status")

head(fData(eset))
```

```{r}
head(pData(eset))
```

```{r}
table(eset$abuse_status)
```

```{r}
eset.log <- log2(exprs(eset))
eset.log[1:5,1:5]
```

```{r}
MED <- apply(eset.log, 1, median)
MAD <- apply(eset.log, 1, mad)
mat <- cbind(MED, MAD)
mat1 <- mat[order(mat[,2], decreasing = TRUE),][1:5000,]
eset.log.filtered <- eset.log[rownames(mat1),]
plot(MED,MAD,pch=20)
points(mat1, pch=20, col='red')
legend(x = "topright", legend = c("top 5k genes by MAD"), col = c('red'), pch = c(20))
```

```{r}
dim(eset.log.filtered)
```

Dimensionality Reduction:

```{r}
library(umap)
eset.log.filtered <- eset.log.filtered[!duplicated(eset.log.filtered), ]  # remove duplicates
ump <- umap(t(eset.log.filtered), n_neighbors = 7, random_state = 123)
plot(ump$layout, main="UMAP plot, nbrs=7", xlab="", ylab="", pch=20, cex=1.5)
library("maptools")  # point labels without overlaps
pointLabel(ump$layout, labels = rownames(ump$layout), method="SANN", cex=0.6)
```

```{r}
dim(eset.log)
```

THIS IS WRONG I DONT EVEN UNDERSTAND WHATS HAPPENING AHHHHHHHHH

```{r}
design1 <- model.matrix(~ eset$abuse_status)
colnames(design1) <- c("NO_ABUSE","ABUSE")
fit1 <- lmFit(eset.log, design1)
fit1 <- eBayes(fit1)
limmaRes1 <- topTable(fit1, coef = "ABUSE", adjust.method="BH", n=Inf, sort.by="P")
nrow(limmaRes1)
```

```{r}
head(limmaRes1[which(limmaRes1$P.Value < .05),])
```
