---
title: "Homework 3"
author: "Mitchell Henschel"
date:   "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: united
    code_folding: show
    toc: true
    style: BS831.css 
    toc_float: true
---

## Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(ggplot2)
library(BS831)
library(ComplexHeatmap)
```

## Simulate a Gene and  Phenotype
```{r}
N <- 100
set.seed(123)

pheno <- sample(0:1, size=N, replace=TRUE)
print(table(pheno))
```

```{r}
## generate a random gene with expression associated to the phenotype
gene <- pheno + rnorm(100,sd=2)

## plot the distribution of the gene within the two phenotype classes
DF <- data.frame(gene=gene,pheno=as.factor(pheno))
p <- ggplot(DF, aes(x=pheno, y=gene, color=pheno)) +
    geom_boxplot()
p
```

## Diffanal by T-Test
```{r}
t1 <- t.test(gene ~ pheno)
print(t1)
```

# Exercise 1
## Diffanal By Permutation Test
Write a function to evaluate the statistical significance of the difference based on a permutation test.
```{r}
permutation.test <- function
(
    gene, 
    pheno, # (binary factor) phenotype
    nperm, # permutation iterations
    alternative=c("two.sided","less","greater"),
    smooth=0, # smoothing factor to avoid pval=0
    seed=NULL, # random seed (for reproducible results)
    do.plot=FALSE # plot the empirical distribution of permuted t-scores
)
{
    ## compute the _observed_ t-score
    obs <- t.test(gene ~ pheno)$statistic[[1]]

    ## compute nperm _permuted_ t scores
    P <- length(gene)
    perm <- c()
    
    for(i in 1:nperm) {
      permed_gene <- sample(gene, size = P, replace = FALSE)
      perm[i] <- t.test(permed_gene ~ pheno)$statistic[[1]]
    }
    
    ## compute the empirical p-value
    pval <- sum(abs(perm) >= abs(obs)) / length(perm)

    if ( do.plot ) {
      ## plot the empirical distribution
      hist(perm, main="Empirical distribution")
      abline(v=obs, lwd=3, col="red")
    }
    return(pval)
}
```

Use the above function to test the simulated gene and phenotype above:
```{r}
## let's record execution time, to see how efficient our algorithm is
time1 <- system.time(
  p2 <- permutation.test(gene=gene,pheno=as.factor(pheno),nperm=1000,seed=123,do.plot=TRUE)
)
```

```{r}
print(time1)
```

```{r}
print(p2)
```

# Exercise 2
## Evaluating the "Soundness" of the Permuation Test
```{r}
## Start by generating random data
N <- 100 # number of samples
M <- 200 # number of "genes"

set.seed(456) # for reproducible results
## generate a random binary phenotype
pheno <- as.factor(base::sample(c("head","tail"),size=N,replace=TRUE))
print(table(pheno))
```

```{r}
## generate a M x N matrix of (gene-by-sample) w/ no differential signal
genes <- matrix(rnorm(M*N),nrow=M,ncol=N)

## let us sort phenotype and genes by phenotype (not necessary)
ord <- order(pheno)
pheno <- pheno[ord]
genes <- genes[,ord]
```

```{r}
## show the heatmap
ha.t <- HeatmapAnnotation(pheno=pheno)

Heatmap(genes,
        name="expression", 
        col=colGradient(c("blue","white","red"),length=13),
        top_annotation=ha.t, 
        cluster_rows=TRUE,
        cluster_columns=FALSE,
        clustering_distance_rows="pearson",
        clustering_method_rows="ward.D",    
        row_title="",
        show_column_names=FALSE,
        show_row_names=FALSE)
```

```{r}
nperm <- 500
## test each gene against the phenotype
time2 <- system.time(
  ptests <- apply(genes,1,permutation.test,pheno=pheno,nperm=nperm)
)
print(time2)        # show execution time
```

```{r}
print(head(ptests)) # show few p-values
```

```{r}
## plot the histogram of the p-values (not shown)
qplot(ptests, geom="histogram")
```

```{r}
## compare to proper distribution
qqplot(x=qnorm(seq(0,1,.01)),y=ptests,xlab="theoretical quantiles",ylab="observed quantiles")
qqline(ptests,distribution=qnorm)
```

# Exercise 3 (Extra Credit)
## Multiple Tests
No.