---
title: "Homework Template"
author: "First Last"
date:   "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: united
    code_folding: show
    toc: true
    style: BS831.css  ## feel free to remove this if you don't want it
    toc_float: true
---

```{r setup, include=FALSE}
## include these options to avoid unnecessarily verbose output
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

You are required to use this template for your homework assignments, or the closest equivalent if you are using Quarto. In particular, each exercise should start with a level 1 heading:

# Exercise X

and if it has multiple questions, each answer should start with a level 2 heading:

## Exercise X.y

Also, please adopt the above preamble (with floating TOC, etc.). Finally, please make sure to knit your document

Adopting these guidelines, will help us in the review of your homework. Please notice that failure to adopt this format will lead to a rejection of the homework.

## Settings

```{r settings, warning=FALSE, message=FALSE}
library(Biobase)
library(BS831)
## or, if failed to install BS831, and cloned it instead, use:
## source("/path/to/github/BS831/clone/R/<needed function>.R")

OMPATH <- Sys.getenv("OMPATH")
## or, if unable to set the OMPATH environment variable in .Renviron, use:
## OMPATH <- "/path/to/Google Drive/BS831 folder/"
```

# Exercise 1

```{r}
```

## Exercise 1.1

## Exercise 1.2

...

# Exercise N

## Exercise N.1

...

## Exercise N.m
