---
title: "Homework 4"
author: "Mitchell Henschel"
date:   "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: united
    code_folding: show
    toc: true
    style: BS831.css 
    toc_float: true
---

## Settings

```{r settings, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(Biobase)
library(dplyr)
library(BS831)
OMPATH <- Sys.getenv("DATAPATH")
```

## Loading the Data

```{r}
DAT <- readRDS(file.path(OMPATH,"HNSC_htseq_normalized_G1vsG3.RDS"))
dim(DAT)
```

```{r}
print(table(DAT$grade,DAT$stage))
```

```{r}
head(fData(DAT))
```

```{r}
## use hypeR helper function
HALLMARK <- hypeR::msigdb_gsets(species="Homo sapiens", category="H")
print(HALLMARK) # show few genesets (use names(HALLMARK$geneset) to see all)
```

```{r}
## extract the relevant geneset
gs <- HALLMARK$genesets[["HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION"]]
length(gs) # count the number of genes
```

```{r}
## see how many genes in the geneset are included in the dataset
length(intersect(gs,fData(DAT)$hgnc_symbol)) 
```

# Exercise 1: Testing for Enrichment

Following the examples and guidelines given in vignettes/docs/KS_enrichment, perform the enrichment test.

i.  First, establish a sorting of the genes in the dataset based on the 'hi (g3) vs. low (g1) grade' phenotype.

```{r}
# Function to extract t-statistics and p-value
tWrapper <- function(gene, pheno) {
    unlist(t.test(gene ~ pheno, var.equal = T)[c("statistic", "p.value")])
}

t <- data.frame(t(apply(exprs(DAT), 1, tWrapper, pheno= DAT$grade))) %>%
    mutate(hgnc_symbol = fData(DAT)$hgnc_symbol)

colnames(t) <- c("t.score", "p.value", "hgnc_symbol")

head(t[,c("hgnc_symbol", "t.score")])
```

ii. Then, determine the ranking of the genes in the geneset (gs) with respect to the sorting just calculated.

```{r}
t <- na.omit(t)
gene_ranks <- t[order(t$t.score),c("hgnc_symbol", "t.score")] 
gene_ranks$rank <- round(rank(gene_ranks$t.score),0)

gs_intersect <- intersect(gs,fData(DAT)$hgnc_symbol)

gene_ranked <- gene_ranks[gene_ranks$hgnc_symbol %in% gs_intersect,] 

head(gene_ranked[order(-gene_ranked$rank),c("hgnc_symbol","rank")])

gene_ranked %>% filter(gene_ranked$hgnc_symbol %in% c("TIMP1", "MATN2", "THBS1", "GREM1", "TFPI2", "CTHRC1"))
```

ii. Finally, use ksGenescore to test for enrichment of the geneset with respect to the phenotype 'hi (g3) vs. low (g1) grade'.

```{r}
# extract the t score
y <- gene_ranked$rank
ksGenescore(34422,y,do.plot=TRUE)
```
